---
title: "Kebbi State — Malaria RDT Test Positivity Rate"
subtitle: "MG2 facility-level analysis vs. DHIS2 indicator"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

```{r paths, include = FALSE}
# ── Paths ──────────────────────────────────────────────────────────────────
# MG2 package functions (load from source during development)
devtools::load_all(
  here::here(), # MG2 project root
  quiet = TRUE
)

nigeria_dir <- "~/Library/Mobile Documents/com~apple~CloudDocs/Projects/Formulas/Nigeria/"
nigeria_file <- "malaria_All-levels_6yrs_2026-05-26.fst"

# Kebbi-only cache — built once from Nigeria, reused for all subsequent steps
kebbi_dir <- "~/Library/Mobile Documents/com~apple~CloudDocs/Projects/MG2/reports/Nigeria/"
kebbi_file <- "kebbi_cache.rds"

# Nigeria DHIS2 credentials
creds <- readRDS(file.path(nigeria_dir, "credentials.rds"))
```

```{r load-packages, include = FALSE}
library(data.table)
library(tsibble)
library(ggplot2)
library(scales)
library(dplyr)
```

---

## 1  Overview

This document computes malaria **RDT test positivity rate (TPR)** for
**Kebbi State, Nigeria** using facility-level data from DHIS2, and compares
the result against the DHIS2 pre-computed indicator.

**Data:** Nigeria DHIS2 national HMIS (`dhis2nigeria.org.ng`), 6-year download,
facility and ward-level records, January 2020 – April 2026.

**TPR definition (RDT only):**

$$
\text{TPR}_{\text{RDT}} = \frac{
  \sum \text{Persons tested positive for malaria by RDT (all age groups)}
}{
  \sum \text{Persons presenting with fever \& tested by RDT (all age groups)}
}
$$

All analyses start from **original (uncleaned) values** (`na_rm = TRUE`).
The final section notes how MAD-based outlier censoring affects results.

---

## 2  Load and filter data

```{r load-data}
kebbi_path <- file.path(kebbi_dir, kebbi_file)

if (!file.exists(kebbi_path)) {
  message("Building Kebbi cache — loading full Nigeria file (~66M rows)...")
  d_all <- fst::read_fst(
    file.path(nigeria_dir, nigeria_file),
    as.data.table = TRUE
  )
  kebbi <- d_all[State == "ke Kebbi State"]
  rm(d_all)
  gc()

  # Month is stored as days-since-1970-01-01 in this FST; convert to yearmonth
  # so that build_combined_dataset() and as.Date(yearmonth(...)) work correctly
  kebbi[, Month := tsibble::yearmonth(as.Date(Month, origin = "1970-01-01"))]

  saveRDS(kebbi, kebbi_path)
  message("Cached: ", kebbi_path)
} else {
  kebbi <- readRDS(kebbi_path)
}

cat(
  "Kebbi facility-months (all elements):",
  format(nrow(kebbi), big.mark = ","),
  "\n"
)
cat(
  "Date range:",
  format(min(as.Date(kebbi$Month)), "%Y-%m"),
  "to",
  format(max(as.Date(kebbi$Month)), "%Y-%m"),
  "\n"
)
cat("Org unit levels:", paste(sort(unique(kebbi$level)), collapse = ", "), "\n")
cat(
  "Facilities (leaf):",
  format(sum(kebbi$effectiveLeaf & !duplicated(kebbi$orgUnit)), big.mark = ","),
  "\n"
)
```

---

## 3  Build TPR combined dataset

MG2's `build_combined_dataset()` sums the four age-group categories of each
data element to produce one numerator and one denominator value per
facility-month, then applies the ratio validity rules:

- **Missing numerator + valid denominator** → numerator treated as 0
  (zero positives is plausible; blank entry is common practice)
- **Present numerator + missing/zero denominator** → excluded
- **Numerator > denominator** → excluded (impossible ratio)

```{r build-tpr}
steps <- list(list(
  operation = "Ratio",
  output_name = "TPR_RDT",
  max_value = 1,
  na_rm = TRUE,
  numerator = list(
    source_file = kebbi_file,
    cleaning_level = "original",
    data_values = c(
      "Persons tested positive for malaria by RDT_<5yrs",
      "Persons tested positive for malaria by RDT_>=5years",
      "Persons tested positive for malaria by RDT_\u22655yrs (excl PW)",
      "Persons tested positive for malaria by RDT_Preg Women (PW)"
    )
  ),
  denominator = list(
    source_file = kebbi_file,
    cleaning_level = "original",
    data_values = c(
      "Persons presenting with fever & tested by RDT_<5yrs",
      "Persons presenting with fever & tested by RDT_>=5years",
      "Persons presenting with fever & tested by RDT_\u22655yrs (excl PW)",
      "Persons presenting with fever & tested by RDT_Preg Women (PW)"
    )
  )
))

tpr_kebbi <- build_combined_dataset(
  steps = steps,
  data_folder = kebbi_dir,
  formula_name = "TPR_RDT",
  verbose = FALSE
)
setDT(tpr_kebbi)
cat("Kebbi TPR rows:", format(nrow(tpr_kebbi), big.mark = ","), "\n")
```

### 3.1  Reporting completeness by category

```{r reporting-breakdown}
fac <- tpr_kebbi[effectiveLeaf == TRUE]

fac[, .(
  total_rows = .N,
  valid_den_missing_num = sum(
    is.na(numerator) & !is.na(denominator) & denominator > 0
  ),
  both_valid = sum(
    !is.na(numerator) &
      !is.na(denominator) &
      denominator > 0 &
      numerator <= denominator
  ),
  orphaned_num = sum(
    !is.na(numerator) & (is.na(denominator) | denominator == 0)
  ),
  impossible_ratio = sum(
    !is.na(numerator) & !is.na(denominator) & numerator > denominator
  ),
  both_missing = sum(is.na(numerator) & is.na(denominator))
)] |>
  t() |>
  as.data.frame() |>
  setNames("n") |>
  within(pct <- round(n / n[1] * 100, 1))
```

### 3.5  Why MG2 counts differ from DHIS2 Indicator A

The raw facility-level sums (before any validity filtering) match DHIS2 Indicator A
(`l2eIpdluuTI`) exactly, every month. The differences visible in Section 6.3 are entirely
from MG2's **facility-level validity rules**, which DHIS2's indicator does not apply.

```{r validity-breakdown}
NUM_VALS <- c(
  "Persons tested positive for malaria by RDT_<5yrs",
  "Persons tested positive for malaria by RDT_>=5years",
  "Persons tested positive for malaria by RDT_≥5yrs (excl PW)",
  "Persons tested positive for malaria by RDT_Preg Women (PW)"
)
DEN_VALS <- c(
  "Persons presenting with fever & tested by RDT_<5yrs",
  "Persons presenting with fever & tested by RDT_>=5years",
  "Persons presenting with fever & tested by RDT_≥5yrs (excl PW)",
  "Persons presenting with fever & tested by RDT_Preg Women (PW)"
)

rdt_raw <- kebbi[effectiveLeaf == TRUE & data %in% c(NUM_VALS, DEN_VALS),
                 .(orgUnit, Month, data, original)]
rdt_raw[, component := fifelse(data %in% NUM_VALS, "numerator", "denominator")]

fac_sums2 <- rdt_raw[, .(
  value = if (all(is.na(original))) NA_real_ else sum(original, na.rm = TRUE)
), by = .(orgUnit, Month, component)]
fac2 <- dcast(fac_sums2, orgUnit + Month ~ component, value.var = "value")

total_fm      <- nrow(fac2)
missing_den   <- fac2[is.na(denominator) | denominator == 0, .N]
num_gt_den    <- fac2[!is.na(numerator) & !is.na(denominator) &
                        denominator > 0 & numerator > denominator, .N]
missing_num   <- fac2[is.na(numerator) & !is.na(denominator) & denominator > 0, .N]

cat("=== Facility-month validity breakdown ===\n")
cat(sprintf("Total facility-months:                %6d\n", total_fm))
cat(sprintf("Missing/zero denominator (excluded):  %6d  (%4.1f%%)\n",
            missing_den,  missing_den  / total_fm * 100))
cat(sprintf("Numerator > denominator (excluded):   %6d  (%4.1f%%)\n",
            num_gt_den,   num_gt_den   / total_fm * 100))
cat(sprintf("Missing numerator, valid den (→ 0):   %6d  (%4.1f%%)\n",
            missing_num,  missing_num  / total_fm * 100))
```

```{r validity-monthly-diff}
# Load DHIS2 Indicator A from cache (full fetch is in Section 6)
d2a_cache <- readRDS(file.path(kebbi_dir, "kebbi_dhis2_indicators.rds"))$correct

fac2[, num_adj := fifelse(
  is.na(numerator) & !is.na(denominator) & denominator > 0, 0, numerator
)]
fac2[, valid := !is.na(num_adj) & !is.na(denominator) &
                  denominator > 0 & num_adj <= denominator]

monthly_raw2 <- fac2[, .(
  raw_num = sum(numerator, na.rm = TRUE),
  raw_den = sum(denominator, na.rm = TRUE)
), by = Month]
monthly_filt2 <- fac2[valid == TRUE, .(
  filt_num = sum(num_adj),
  filt_den = sum(denominator)
), by = Month]

monthly_raw2[,  date := as.Date(yearmonth(as.character(Month)))]
monthly_filt2[, date := as.Date(yearmonth(as.character(Month)))]

diff_tbl <- merge(
  monthly_raw2[,  .(date, raw_num,  raw_den)],
  monthly_filt2[, .(date, filt_num, filt_den)],
  by = "date"
)
diff_tbl <- merge(
  diff_tbl,
  as.data.table(d2a_cache)[, .(date, d2_num = numerator, d2_den = denominator)],
  by = "date"
)
diff_tbl[, `:=`(num_diff = raw_num - filt_num, den_diff = raw_den - filt_den)]

cat("=== Raw sums match DHIS2 Indicator A exactly? ===\n")
cat("Months where raw_num != d2_num:", diff_tbl[raw_num != d2_num, .N], "\n")
cat("Months where raw_den != d2_den:", diff_tbl[raw_den != d2_den, .N], "\n\n")

cat("=== Months where MG2 filtered != raw (i.e., differs from DHIS2) ===\n")
diff_tbl[num_diff != 0 | den_diff != 0,
         .(date,
           `DHIS2 A (=raw) num` = format(d2_num,   big.mark = ","),
           `MG2 filt num`       = format(filt_num,  big.mark = ","),
           `num removed`        = format(num_diff,  big.mark = ","),
           `DHIS2 A (=raw) den` = format(d2_den,   big.mark = ","),
           `MG2 filt den`       = format(filt_den,  big.mark = ","),
           `den removed`        = format(den_diff,  big.mark = ","))]
```

The raw facility sums equal DHIS2 Indicator A to the digit — confirming the MG2 download
covers identical facilities and data. Every difference between the MG2 TPR series and
Indicator A comes from three validity rules applied at facility level:

1. **Numerator > denominator — primary driver.** Facility-months where recorded positives
   exceed tested count are dropped as impossible. Because the numerator was inflated *above*
   the denominator in those rows, removing them reduces the numerator more than the
   denominator — explaining why `num removed` is consistently larger than `den removed`.

2. **Missing/zero denominator.** Facility-months with no tested count are excluded entirely.
   DHIS2 sums those numerator values regardless.

3. **Missing numerator → 0 (minor, offsetting).** When a facility has a valid tested count
   but no positives recorded, MG2 imputes zero positives. This slightly *raises* the MG2
   numerator relative to a strict NA-drop, but the effect (0.4% of facility-months) is
   smaller than the exclusions above.

The magnitude of removed counts has grown in recent years (2024–2025: up to ~9,000
positives removed in a single month), suggesting a deteriorating data quality signal —
more facility-months with impossible `num > den` entries — worth monitoring independently
of the TPR calculation itself.

---

## 4  TPR trend — Kebbi State

### 4.1  All facilities (aggregate)

```{r kebbi-national-tpr}
# Apply validity rules and aggregate monthly
kebbi_monthly <- fac[
  !is.na(denominator) &
    denominator > 0 &
    (is.na(numerator) | numerator <= denominator),
  .(
    num = sum(fifelse(is.na(numerator), 0, numerator)),
    den = sum(denominator),
    n_fac = .N
  ),
  by = Month
]
kebbi_monthly[, `:=`(
  tpr_pct = num / den * 100,
  date = as.Date(yearmonth(as.character(Month)))
)]
setorder(kebbi_monthly, date)

ggplot(kebbi_monthly, aes(x = date)) +
  geom_line(aes(y = tpr_pct), colour = "#1f78b4", linewidth = 0.9) +
  geom_point(aes(y = tpr_pct, size = n_fac), colour = "#1f78b4", alpha = 0.7) +
  scale_size_continuous(name = "Facilities\nreporting", range = c(1, 4)) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b\n%Y") +
  labs(
    title = "Kebbi State — RDT Test Positivity Rate (all facilities)",
    subtitle = "Original values, na_rm = TRUE. Point size = facilities with valid denominator.",
    x = NULL,
    y = "TPR (%)"
  ) +
  theme_bw()
```

```{r kebbi-tpr-table}
kebbi_monthly[, .(
  date,
  positives = num,
  tested = den,
  tpr_pct = round(tpr_pct, 1),
  n_fac
)]
```

### 4.2  Facility-level TPR distribution

```{r facility-tpr-dist}
# Compute per-facility average TPR for recent 2 years
recent_cutoff <- as.Date(max(kebbi_monthly$date)) - 365 * 2

fac_avg <- fac[
  as.Date(yearmonth(as.character(Month))) >= recent_cutoff &
    !is.na(denominator) &
    denominator > 0 &
    (is.na(numerator) | numerator <= denominator),
  .(
    num_total = sum(fifelse(is.na(numerator), 0, numerator)),
    den_total = sum(denominator),
    n_months = .N
  ),
  by = orgUnit
]
fac_avg[, tpr_pct := num_total / den_total * 100]
fac_avg <- fac_avg[den_total > 0]

# Join facility names from kebbi (orgUnitName not preserved by build_combined_dataset)
ou_names <- unique(kebbi[effectiveLeaf == TRUE, .(orgUnit, orgUnitName)])
fac_avg <- ou_names[fac_avg, on = "orgUnit"]

ggplot(fac_avg, aes(x = tpr_pct)) +
  geom_histogram(
    binwidth = 5,
    fill = "#1f78b4",
    colour = "white",
    alpha = 0.8
  ) +
  geom_vline(
    xintercept = kebbi_monthly[
      date >= recent_cutoff,
      sum(num) / sum(den) * 100
    ],
    linetype = "dashed",
    colour = "#e31a1c",
    linewidth = 0.9
  ) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = paste0(
      "Facility TPR distribution — last 2 years (n = ",
      nrow(fac_avg),
      " facilities)"
    ),
    subtitle = "Red dashed line = aggregate Kebbi TPR for same period",
    x = "Facility TPR (%)",
    y = "Count"
  ) +
  theme_bw()
```

---

## 5  Champion facility analysis

Champion facilities are those that report consistently — here defined as
having a valid denominator (tested count > 0) in at least a specified
fraction of months during the analysis window.

```{r champion-definition}
# Analysis window: full date range in the data
date_range <- range(kebbi_monthly$date)
n_months_total <- nrow(kebbi_monthly)

# Count months each facility reported a valid denominator
fac_report_counts <- fac[
  !is.na(denominator) & denominator > 0,
  .(n_reported = .N),
  by = orgUnit
]
fac_all <- fac[, .(n_total = .N), by = orgUnit]
fac_summary <- merge(fac_all, fac_report_counts, by = "orgUnit", all.x = TRUE)
fac_summary[is.na(n_reported), n_reported := 0]
fac_summary[, pct_reported := n_reported / n_months_total * 100]

cat(
  "Facilities with ≥80% reporting:",
  sum(fac_summary$pct_reported >= 80),
  "\n"
)
cat(
  "Facilities with ≥60% reporting:",
  sum(fac_summary$pct_reported >= 60),
  "\n"
)
cat(
  "Facilities with ≥40% reporting:",
  sum(fac_summary$pct_reported >= 40),
  "\n"
)

ggplot(fac_summary, aes(x = pct_reported)) +
  geom_histogram(
    binwidth = 5,
    fill = "#33a02c",
    colour = "white",
    alpha = 0.8
  ) +
  geom_vline(xintercept = 80, linetype = "dashed", colour = "#e31a1c") +
  labs(
    title = "Distribution of facility reporting frequency",
    subtitle = "Red line = 80% threshold (champion criterion)",
    x = "% of months with valid denominator",
    y = "Facilities"
  ) +
  theme_bw()
```

```{r champion-tpr}
champion_ids <- fac_summary[pct_reported >= 80, orgUnit]
cat("Champion facilities (≥80%):", length(champion_ids), "\n")

# Monthly TPR: champions vs. all
champ_monthly <- fac[
  orgUnit %in%
    champion_ids &
    !is.na(denominator) &
    denominator > 0 &
    (is.na(numerator) | numerator <= denominator),
  .(
    num = sum(fifelse(is.na(numerator), 0, numerator)),
    den = sum(denominator),
    n_fac = .N
  ),
  by = Month
]
champ_monthly[, `:=`(
  tpr_pct = num / den * 100,
  date = as.Date(yearmonth(as.character(Month))),
  group = "Champions (≥80% reporting)"
)]

all_monthly <- copy(kebbi_monthly)[, group := "All facilities"]
setnames(all_monthly, "tpr_pct", "tpr_pct")

plot_champ <- rbind(
  champ_monthly[, .(date, tpr_pct, n_fac, group)],
  all_monthly[, .(date, tpr_pct, n_fac, group)]
)

ggplot(plot_champ, aes(x = date, y = tpr_pct, colour = group)) +
  geom_line(linewidth = 0.9) +
  geom_point(aes(size = n_fac), alpha = 0.7) +
  scale_colour_manual(
    values = c(
      "Champions (≥80% reporting)" = "#33a02c",
      "All facilities" = "#1f78b4"
    )
  ) +
  scale_size_continuous(name = "Facilities", range = c(1, 4)) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b\n%Y") +
  labs(
    title = "Kebbi State RDT TPR — champions vs. all facilities",
    x = NULL,
    y = "TPR (%)",
    colour = NULL
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

---

## 6  Comparison with DHIS2 indicators

Two DHIS2 indicators cover RDT positivity. They have **different formulas**
and the comparison reveals an important data quality issue.

| Indicator | ID | Formula | Issues |
|-----------|-----|---------|--------|
| `% of Fever cases tested positive with RDT` | `l2eIpdluuTI` (**Indicator A**) | Positives (default total) / Fever cases tested (default total) | Denominator may include untested fever cases → TPR deflated |
| `Test Positivity Rate(TPR) (RDT)` | `uiA07P72s2Z` (**Indicator B**) | Positives (4 groups) / **Positives (total) + Tested (3 groups, new scheme only)** | Positives added to denominator; COC scheme mix; missing new-scheme `<5yrs` |

**Indicator A — `l2eIpdluuTI`** uses default aggregation (DHIS2 sums all category option combos).

*Denominator concern:* the data element "Persons presenting with fever & tested by RDT" (`r6WOvUlcQm6`)
is intended to count only those actually tested. If health workers enter total fever attendees
rather than the tested subset, the denominator is inflated by untested cases — since
they cannot be positive, the numerator is unaffected and reported TPR is deflated.

| Role | Data Element | DE UID | Category | COC UID |
|------|-------------|--------|----------|---------|
| Numerator | Persons tested positive for malaria by RDT | `GEd2F6skCpT` | *(default — all age groups)* | — |
| Denominator | Persons presenting with fever & tested by RDT | `r6WOvUlcQm6` | *(default — all age groups)* | — |

**Indicator B — `uiA07P72s2Z`** explicitly names category option combos, mixing old and new age-disaggregation schemes:

| Role | Data Element | DE UID | Category | COC UID | Scheme |
|------|-------------|--------|----------|---------|--------|
| Numerator | Persons tested positive for malaria by RDT | `GEd2F6skCpT` | `<5yrs` | `W50DgZrPrUk` | Old |
| Numerator | Persons tested positive for malaria by RDT | `GEd2F6skCpT` | `>=5years` | `ItGQGDRFD4t` | Old |
| Numerator | Persons tested positive for malaria by RDT | `GEd2F6skCpT` | `≥5yrs (excl PW)` | `v0QZcUTuyjD` | New |
| Numerator | Persons tested positive for malaria by RDT | `GEd2F6skCpT` | `Preg Women (PW)` | `V86mzY788hR` | New |
| ~~Numerator~~ | ~~Persons tested positive for malaria by RDT~~ | `GEd2F6skCpT` | ~~`<5yrs` (new scheme)~~ | `Bl7e1tsRGoZ` | ~~New~~ **missing** |
| **Denominator** ✗ | **Persons tested positive for malaria by RDT** | `GEd2F6skCpT` | ***(default — all age groups)*** | — | **ERROR: adds positives to denominator** |
| Denominator | Persons presenting with fever & tested by RDT | `r6WOvUlcQm6` | `<5yrs` | `Bl7e1tsRGoZ` | New |
| Denominator | Persons presenting with fever & tested by RDT | `r6WOvUlcQm6` | `≥5yrs (excl PW)` | `v0QZcUTuyjD` | New |
| Denominator | Persons presenting with fever & tested by RDT | `r6WOvUlcQm6` | `Preg Women (PW)` | `V86mzY788hR` | New |
| ~~Denominator~~ | ~~Persons presenting with fever & tested by RDT~~ | `r6WOvUlcQm6` | ~~`<5yrs` + `>=5years` (old scheme)~~ | — | ~~Old~~ **missing** |

Two errors compound: (1) the denominator adds the **default-aggregation total of RDT positives**
(`#{GEd2F6skCpT}` without a COC — DHIS2 interprets this as the sum across all category option
combos), inflating the denominator when positivity is high and deflating apparent TPR; (2) the
numerator omits the new-scheme `<5yrs` positives (`Bl7e1tsRGoZ`), and neither numerator nor
denominator covers old-scheme tested (`<5yrs`, `>=5years` of `r6WOvUlcQm6`) since that data
element does not have the `Age(Malaria)` category combination.

**Pre-transition (falsely elevated, then suppressed).** Before the NH19 age-scheme migration,
new-scheme tested categories for `r6WOvUlcQm6` returned zero. Indicator B's denominator
therefore collapsed to `#{GEd2F6skCpT}` default (positives only). With numerator ≈ denominator,
Indicator B would return TPR ≈ 100% — which DHIS2 suppresses for percentage-type indicators.
This explains why Indicator B returns **no rows for 2020–2021** in the Kebbi data: not absent
testing, but an implicit value >100% that DHIS2 will not return. The post-transition deflation
(~26%) and the pre-transition inflation (~100%) are both consequences of the same formula errors.

### 6.1  Which indicator do programs and implementing partners use?

Nigeria's National Health Management Information System (NHMIS) — deployed on DHIS2 — is the
official source of malaria surveillance data for the National Malaria Elimination Programme
(NMEP), the US President's Malaria Initiative (PMI), and implementing partners. The NMEP's
2021–2025 National Malaria Strategic Plan names HMIS-derived test positivity rate as a core
programme indicator (NMEP, 2021). PMI's FY 2023 Nigeria Malaria Operational Plan (USAID, 2023)
and the PMI for States (PMI-S/MSH) programme cite DHIS2 TPR values in all 11 focus states,
including Kebbi (~80%) and Zamfara (62–89%), to guide resource allocation (Data for Impact,
2023). Malaria Consortium's SMC implementation in Kebbi, Sokoto, and Bauchi similarly relies on
HMIS malaria testing data for monitoring and evaluation (GiveWell, 2023).

**No publicly available programme report, operational plan, or implementing-partner document names
the specific DHIS2 indicator UID or formula used to derive TPR.** All sources describe the metric
generically as "positive tests / total tests performed among febrile patients" (e.g., Dahunsi
et al., 2025; Oladoyin et al., 2025), without citing a DHIS2 indicator name or ID. The practical
consequence is significant: the two available indicators carry different names in the DHIS2
interface — Indicator B: "Test Positivity Rate(TPR) (RDT)" (`uiA07P72s2Z`); Indicator A:
"% of Fever cases tested positive with RDT" (`l2eIpdluuTI`). An analyst searching for "TPR"
would encounter Indicator B (`uiA07P72s2Z`) first; Indicator A (`l2eIpdluuTI`) uses different
terminology that does not include the "TPR" acronym.

### 6.2  Literature citing Nigeria HMIS/DHIS2 TPR without specifying the indicator

A consistent pattern across the recent literature on Nigeria malaria is that authors extract
test positivity data from DHIS2 without documenting which of the two indicators — or which
underlying data elements — were used:

- **Dahunsi et al. (2025)** extracted RDT-based TPR for Oyo State from DHIS2 for 2021–2022
  (TPR 70–82%), reporting a pre-post decline attributed to RDT quality improvement. The paper
  defines TPR as "positive tests as a proportion of total tests performed among febrile patients"
  but cites no DHIS2 indicator name or UID. Funded by PMI; implemented by MSH.
  *Malaria Journal* 24:65. ([PMC12016084](https://pmc.ncbi.nlm.nih.gov/articles/PMC12016084/))

- **Oladoyin et al. (2025)** selected study facilities in Sokoto and Oyo States on the basis of
  "the highest TPR reported on the DHIS2," defining TPR as "number of positive tests divided by
  total number of tests performed." No indicator name or UID is given. Funded by PMI Insights
  and the Bill & Melinda Gates Foundation.
  *Malaria Journal* 24 (2025). ([PMC12595662](https://pmc.ncbi.nlm.nih.gov/articles/PMC12595662/))

- **Gnansounou et al. (2025)** assessed concordance of RDT results between facility registers
  and DHIS2 across Benin, Côte d'Ivoire, Nigeria, and Uganda, extracting "number of positive
  RDT results and total RDTs administered" from DHIS2. No indicator name or UID specified for
  any country. PMI-funded.
  *Malaria Journal* 24 (2025). ([PMC12219610](https://pmc.ncbi.nlm.nih.gov/articles/PMC12219610/))

- **Data for Impact / D4I (2023)** evaluated DHIS2 data quality in Kebbi, Zamfara, and Ebonyi
  States, reporting that "malaria test positivity has been relatively stable at around 80% in
  primary health care facilities" in Kebbi and 62–89% in Zamfara. No indicator name cited.
  *Nigeria DHIS2 Evaluation Baseline Results*, Data for Impact Project (USAID).

- **Multi-country concordance study (2025)** — Nigeria, Uganda, Benin, Côte d'Ivoire — assessed
  agreement between DHIS2 aggregate data and source registers. TPR is used as a selection
  criterion but neither defined nor attributed to a specific DHIS2 indicator.
  ([PMC12866365](https://pmc.ncbi.nlm.nih.gov/articles/PMC12866365/))

The gap matters: if any of these studies extracted data through Indicator B (`uiA07P72s2Z`)
— whether by selecting it from a DHIS2 dashboard or via an analytics API call referencing it —
the TPR values reported would be systematically deflated (denominator inflated by adding total
positives), and the direction and magnitude of any intervention effect or state-level comparison
would be affected accordingly. None of the papers provide sufficient methodological detail to
rule this out.

```{r dhis2-fetch-setup}
library(httr)

DHIS2_CACHE <- file.path(kebbi_dir, "kebbi_dhis2_indicators.rds")

kebbi_ou <- "M689V9w3Gs3"

# One period string per year — sent as separate requests to avoid timeouts.
# Each element is a ";"-joined list of YYYYMM codes for that calendar year.
period_batches <- c(
  paste0(2020, sprintf("%02d", 1:12), collapse = ";"),
  paste0(2021, sprintf("%02d", 1:12), collapse = ";"),
  paste0(2022, sprintf("%02d", 1:12), collapse = ";"),
  paste0(2023, sprintf("%02d", 1:12), collapse = ";"),
  paste0(2024, sprintf("%02d", 1:12), collapse = ";"),
  paste0(2025, sprintf("%02d", 1:12), collapse = ";"),
  paste0(2026, sprintf("%02d", 1:4),  collapse = ";")
)

# fetch_indicator(): loops over annual period batches, one request per year,
# then combines results.  Returns NULL only if every batch fails.
fetch_indicator <- function(ind_id, label, base_url, username, password,
                            batch_pause = 1) {
  cat("Fetching:", label, "\n")
  cat("Indicator UID:", ind_id, "| batches:", length(period_batches), "\n\n")

  batch_results <- vector("list", length(period_batches))

  for (i in seq_along(period_batches)) {
    yr_label <- substr(period_batches[[i]], 1, 4)   # "2020", "2021", …
    url <- paste0(
      base_url,
      "api/analytics.json",
      "?dimension=dx:", ind_id,
      "&dimension=pe:", period_batches[[i]],
      "&dimension=ou:", kebbi_ou,
      "&includeNumDen=true&skipRounding=false&displayProperty=NAME"
    )

    cat("  Batch", i, "(", yr_label, "):", url, "\n")

    resp <- tryCatch(
      GET(url, authenticate(username, password), timeout(30)),
      error = function(e) {
        cat("    ERROR (connection):", conditionMessage(e), "\n")
        NULL
      }
    )

    if (is.null(resp)) {
      cat("    RESULT: no response (timeout or DNS failure) — skipping\n")
      next
    }

    sc <- status_code(resp)
    cat("    HTTP:", sc)

    if (sc == 401) {
      cat(" — authentication failed, aborting all batches\n")
      return(NULL)
    }
    if (sc != 200) {
      cat(" — unexpected status, skipping batch\n")
      cat("    Body:", rawToChar(content(resp, "raw")), "\n")
      next
    }

    d <- content(resp, as = "parsed")
    if (length(d$rows) == 0) {
      cat(" — 0 rows returned\n")
      next
    }

    rows <- lapply(d$rows, function(r) {
      data.frame(
        period      = r[[2]],
        value       = as.numeric(r[[4]]),
        numerator   = as.numeric(r[[5]]),
        denominator = as.numeric(r[[6]]),
        stringsAsFactors = FALSE
      )
    })
    batch_results[[i]] <- as.data.table(do.call(rbind, rows))
    cat(" —", nrow(batch_results[[i]]), "rows\n")

    if (i < length(period_batches)) Sys.sleep(batch_pause)
  }

  combined <- rbindlist(Filter(Negate(is.null), batch_results))
  if (nrow(combined) == 0) {
    cat("\n  RESULT: no data returned across all batches\n")
    return(NULL)
  }

  combined[, `:=`(
    date  = as.Date(paste0(period, "01"), format = "%Y%m%d"),
    label = label
  )]
  cat("\n  TOTAL:", nrow(combined), "rows,",
      format(min(combined$date), "%Y-%m"), "–",
      format(max(combined$date), "%Y-%m"), "\n")
  combined
}
```

```{r dhis2-fetch, eval = TRUE}
# ── Load from cache or fetch from DHIS2 ────────────────────────────────────
# Re-run this chunk interactively to force a fresh download:
#   delete kebbi_dhis2_indicators.rds, or set force_fetch <- TRUE

force_fetch <- FALSE   # ← set TRUE to bypass cache

if (!force_fetch && file.exists(DHIS2_CACHE)) {

  cat("Loading DHIS2 indicators from cache:", DHIS2_CACHE, "\n")
  dhis2_cache <- readRDS(DHIS2_CACHE)
  dhis2_correct <- dhis2_cache$correct
  dhis2_errored <- dhis2_cache$errored
  cat("  correct indicator rows:", if (is.null(dhis2_correct)) "NULL" else nrow(dhis2_correct), "\n")
  cat("  errored indicator rows:", if (is.null(dhis2_errored)) "NULL" else nrow(dhis2_errored), "\n")

} else {

  cat("Fetching DHIS2 indicators from live instance...\n")
  cat("Base URL:", creds$url, "\n\n")

  dhis2_correct <- fetch_indicator(
    "l2eIpdluuTI",
    "DHIS2 Indicator A (l2eIpdluuTI)",
    creds$url, creds$username, creds$password
  )

  cat("\n")

  dhis2_errored <- fetch_indicator(
    "uiA07P72s2Z",
    "DHIS2 Indicator B (uiA07P72s2Z)",
    creds$url, creds$username, creds$password
  )

  if (!is.null(dhis2_correct) || !is.null(dhis2_errored)) {
    saveRDS(
      list(correct = dhis2_correct, errored = dhis2_errored),
      DHIS2_CACHE
    )
    cat("\nSaved to cache:", DHIS2_CACHE, "\n")
  } else {
    cat("\nBoth fetches failed — nothing saved to cache.\n")
  }
}

if (is.null(dhis2_correct)) {
  cat("\n⚠  dhis2_correct is NULL — Section 6 comparison plots will be skipped.\n")
}
```

```{r compare-plot, eval = !is.null(dhis2_correct)}
mg2_line <- kebbi_monthly[, .(
  date,
  value = tpr_pct,
  numerator = num,
  denominator = den,
  label = "MG2 (facility-level, original values)"
)]

all_series <- rbindlist(
  Filter(
    Negate(is.null),
    list(
      mg2_line,
      dhis2_correct[, .(date, value, numerator, denominator,
                        label = "DHIS2 Indicator A (l2eIpdluuTI)")],
      dhis2_errored[, .(date, value, numerator, denominator,
                        label = "DHIS2 Indicator B (uiA07P72s2Z)")]
    )
  ),
  fill = TRUE
)

ggplot(all_series, aes(x = date, y = value, colour = label)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5, alpha = 0.8) +
  scale_colour_manual(
    values = c(
      "MG2 (facility-level, original values)" = "#1f78b4",
      "DHIS2 Indicator A (l2eIpdluuTI)"       = "#33a02c",
      "DHIS2 Indicator B (uiA07P72s2Z)"       = "#e31a1c"
    )
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b\n%Y") +
  labs(
    title = "Kebbi State RDT TPR — MG2 vs. DHIS2 indicators",
    subtitle = "Indicator B (red) adds RDT positives to denominator (~50pp gap between indicators)",
    x = NULL,
    y = "TPR (%)",
    colour = NULL
  ) +
  theme_bw() +
  theme(legend.position = "bottom", legend.text = element_text(size = 8))
```

### 6.3  Numerator / denominator comparison

```{r numden-compare, eval = !is.null(dhis2_correct)}
# Compare counts for DHIS2 Indicator A vs. MG2
nd_compare <- merge(
  mg2_line[, .(date, mg2_num = numerator, mg2_den = denominator)],
  dhis2_correct[, .(date, dhis2_num = numerator, dhis2_den = denominator)],
  by = "date",
  all = TRUE
)
nd_compare[, `:=`(
  num_ratio = round(mg2_num / dhis2_num, 2),
  den_ratio = round(mg2_den / dhis2_den, 2)
)]
nd_compare
```

Raw facility sums match DHIS2 Indicator A exactly. Differences here reflect
MG2's validity filtering — see Section 3.5 for a full breakdown.

### 6.4  NHMIS form changes and age-disaggregation transition

Nigeria's NHMIS form was revised from a **2013 version** (dataset UID `lyVV9bPLlVy`) to a
**2019 version** (`O82o1WlMisO`). Both versions include `GEd2F6skCpT` (positives) and
`r6WOvUlcQm6` (tested). The revision changed the malaria testing age-disaggregation scheme:

| Scheme | Category Combo | Groups | Key COC UIDs |
|--------|---------------|--------|-------------|
| **Old** (pre-2019) | `Age(Malaria)` | `<5yrs`, `>=5years` | `W50DgZrPrUk`, `ItGQGDRFD4t` |
| **New** (NH19) | `NH19_Malaria Testing` | `<5yrs`, `≥5yrs (excl PW)`, `Preg Women (PW)` | `Bl7e1tsRGoZ`, `v0QZcUTuyjD`, `V86mzY788hR` |

The "NH19" prefix in the new category combination name corresponds to the 2019 NHMIS revision.
In current metadata, both data elements are assigned `NH19_Malaria Testing`; the `Age(Malaria)`
combo now appears only for community form data elements.

Kebbi State data reveals a sharp, coordinated rollout:

```{r scheme-transition}
pos_long <- kebbi[
  effectiveLeaf == TRUE &
    grepl("Persons tested positive for malaria by RDT", data)
]

pos_long[, scheme := fcase(
  data %like% ">=5years",                  "old",
  data %like% "≥5yrs|Preg Women",     "new",
  default                                  = NA_character_
)]

scheme_facs <- pos_long[
  !is.na(scheme) & !is.na(original),
  .(n_fac = uniqueN(orgUnit)),
  by = .(Month, scheme)
]
scheme_facs[, date := as.Date(Month)]

dcast(
  scheme_facs[date >= as.Date("2020-09-01") & date <= as.Date("2021-06-01")],
  date ~ scheme, value.var = "n_fac", fill = 0L
)
```

Zero facilities reported under the new scheme before January 2021; by February 2021 the old
scheme had zero facilities. This near-simultaneous cutover across hundreds of facilities is
consistent with a coordinated national DHIS2 form rollout.

**Consequences for the indicators.** Indicator B's numerator references both old-scheme COC
UIDs (`W50DgZrPrUk`, `ItGQGDRFD4t`) and new-scheme UIDs; its denominator uses only new-scheme
COC UIDs for `r6WOvUlcQm6`. Pre-transition, new-scheme tested = 0 → TPR → ~100% (suppressed).
Post-transition, old-scheme COC UIDs return no new data; the formula error (positives in the
denominator) produces the observed ~50pp deflation.

An additional scope difference: `GEd2F6skCpT` (positives) also appears in the **Community
Monthly Summary Form**, while `r6WOvUlcQm6` (tested) does not. If community health extension
worker positives are aggregated into state totals without a corresponding tested denominator,
Indicator A's default-aggregation denominator undercounts tested cases from community reporting.

### 6.5  What we cannot answer and what to investigate

The metadata and empirical data reveal *what* the formulas are and *how* they diverge, but
several historical questions cannot be answered from available evidence:

1. **When was Indicator B created, and was the formula ever correct?** The DHIS2 API exposes
   `created` and `lastUpdated` timestamps for indicator objects. If Indicator B predates the
   NH19 transition (≈ 2020), its old-scheme COC UIDs may have been valid references at the time
   — the error arising from category combination changes without corresponding formula updates
   (configuration drift). If created after the transition, the mixing of old and new UIDs is an
   error from inception.

2. **Was `Age(Malaria)` ever assigned to `GEd2F6skCpT` or `r6WOvUlcQm6`?** In current metadata
   these elements carry `NH19_Malaria Testing`. An audit trail of category combination
   assignments — if the Nigeria DHIS2 instance maintains one — would confirm whether they
   previously used `Age(Malaria)`, explaining the old-scheme UIDs in Indicator B's numerator.

3. **Does the Nigeria DHIS2 instance maintain a change log?** Some DHIS2 versions record
   metadata changes. Without this, reconstructing when and why Indicator B's formula was written
   as it is requires human investigation of the original configuration staff.

4. **How consistently is `r6WOvUlcQm6` (tested) recorded?** A facility-level comparison of
   months with valid numerator vs. valid denominator would reveal whether the "tested" field is
   systematically missing for certain facility types — biasing TPR regardless of which indicator
   is used.

5. **Do community records contribute to the state aggregate?** If `GEd2F6skCpT` values from
   the Community form are included in state rollups without a corresponding tested denominator,
   both indicators understate true TPR.

### 6.6  Computing a transparent TPR with MG2

Section 3 of this report demonstrates the alternative to relying on pre-built DHIS2 indicators:
building the ratio directly from facility-level data element values. The `steps` list in the
`build-tpr` chunk is a complete formula specification — data element names, category option
combos, and validity rules — that is explicit, versioned with the analysis, and reproducible
without a live DHIS2 connection.

MG2's Combine module (`build_combined_dataset()` with `operation = "Ratio"`) provides this
capability interactively in the Shiny application. Users can:

- Select specific data element × category combinations for numerator and denominator
- Choose which age-disaggregation scheme applies to their analysis window
- Apply validity rules (missing numerator → 0; numerator > denominator → excluded)
- Save a fully documented formula alongside the computed dataset

This addresses precisely the gap identified in Sections 6.1–6.2: none of the five programme
reports and published studies cited in those sections provided enough formula detail for a reader
to reconstruct the analysis. TPR computed from data elements makes that documentation an
automatic byproduct of the workflow — and makes formula errors like those in Indicator B
detectable before results are published.

---

## 7  Notes on outlier cleaning

This analysis uses **original values** throughout. The same analysis with
`cleaning_level = "mad10"` would exclude values flagged as more than 10×
the median absolute deviation from the series median. To explore this,
change the `cleaning_level` argument in Section 3 to `"mad10"` and re-render.

```{r cleaning-preview}
# Preview: how many facility-months are flagged at each cleaning level in Kebbi
rdt_pos <- kebbi[data %like% "Persons tested positive.*RDT"]
rdt_tested <- kebbi[data %like% "Persons presenting.*RDT"]

cat("=== Numerator (RDT positives) ===\n")
rdt_pos[, .(
  total = .N,
  over_max = sum(over_max, na.rm = TRUE),
  mad15 = sum(mad15, na.rm = TRUE),
  mad10 = sum(mad10, na.rm = TRUE),
  seasonal5 = sum(seasonal5, na.rm = TRUE),
  seasonal3 = sum(seasonal3, na.rm = TRUE)
)]

cat("\n=== Denominator (RDT tested) ===\n")
rdt_tested[, .(
  total = .N,
  over_max = sum(over_max, na.rm = TRUE),
  mad15 = sum(mad15, na.rm = TRUE),
  mad10 = sum(mad10, na.rm = TRUE),
  seasonal5 = sum(seasonal5, na.rm = TRUE),
  seasonal3 = sum(seasonal3, na.rm = TRUE)
)]
```

---

*Analysis run: `r Sys.time()` | Data: `r nigeria_file` | MG2 version: `r packageVersion("MG2")`*
