---
title: "Burden Estimation: Standalone Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Burden Estimation: Standalone Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE   # requires a local data file; set eval = TRUE after loading data
)
```

## Overview

Routine health facility reporting rarely captures all cases. Facilities may not report
every month, and some confirmed cases may be recorded as zero rather than missing.
The MG2 burden estimation module adjusts for these gaps by:

1. Identifying **champion facilities** — those that report consistently and are
   used as the reference population.
2. **Extrapolating** from champion rates to all facilities in each region using
   one of several statistical methods.
3. Producing **confidence intervals** via bootstrap resampling (1,000 draws).

This vignette shows how to run the full burden estimation workflow from an `.fst`
or `.rds` processed dataset, without opening the Shiny application.

---

## 1  Load and prepare data

The starting point is a processed dataset — the output of MG2's Data module,
stored as `.fst` (or `.rds`). Set the path to your file:

```{r load-data}
library(MG2)
library(tsibble)     # yearmonth()
library(data.table)  # setDT(), :=

data_file <- "/path/to/your/malaria_All-levels_3yrs_2026-05-25.fst"

d_raw <- read_file(data_file)
```

```{r clean-data}
# cleanedData() filters to facility-level rows (effectiveLeaf == TRUE) and
# creates the `dataCol` numeric column from the raw `original` values.
# Pass source = "Original" to use unmodified values; use the error= argument
# to apply a specific outlier threshold instead.
d_clean <- cleanedData(d_raw, source = "Original")
```

---

## 2  Define target elements and time period

Burden estimates apply to a specific set of data elements and a date window.
Here we estimate confirmed malaria cases across three age groups for calendar
year 2024 and 2025:

```{r define-targets}
target_elements <- c(
  "MAL - Confirmed malaria cases_0-4 years",
  "MAL - Confirmed malaria cases_5-14 years",
  "MAL - Confirmed malaria cases_15+ years"
)

period_start <- yearmonth("2024 Jan")
period_end   <- yearmonth("2025 Dec")
```

---

## 3  Identify champion facilities

`selectedData()` filters the data to the target elements and identifies champion
facilities — those that reported in at least `n - missing_reports` months during
the period. It adds a `Selected` column (`"Champion"` / `"Non-Champion"`) to
every row.

```{r select-data}
d <- selectedData(
  d_clean,
  levelNames      = c("Country", "Province", "District", "Facility"),
  data_categories = target_elements,
  startingMonth   = period_start,
  endingMonth     = period_end,
  missing_reports = 2L   # allow up to 2 missed months in a 24-month window
)

# How many champion facilities per province?
champ_by_province <- unique(d[d$Selected == "Champion", c("orgUnit", "Province")])
table(champ_by_province$Province)
```

```
01 Vientiane Capital          02 Phongsali          03 Louangnamtha
                   1                     4                        0
        04 Oudomxai             05 Bokeo       06 Louangphabang
                   0                     0                        0
        07 Houaphan         08 Xainyabouli       09 Xiangkhouang
                   2                     4                        0
        10 Vientiane       11 Bolikhamxai         12 Khammouan
                   2                     0                        0
      13 Savannakhet           14 Salavan             15 Xekong
                  30                    18                       19
        16 Champasak           17 Attapu          18 Xaisomboun
                  22                     7                        5
```

Provinces with zero champions will receive burden estimates via national-level
extrapolation in methods A and B, and via the champion-distribution fallback
in C1 and C2.

```{r prep-value}
# The burden functions use a `value` column. Create it from `dataCol`
# (the numeric clean column that selectedData() produces).
setDT(d)
d[, value := dataCol]
```

---

## 4  Run burden estimates

### Method A — Champion multiple

The simplest estimator: take the per-facility total across champion facilities,
then scale up by the ratio of all facilities to champion facilities in each
region. Non-champion facility totals are predicted by resampling from the
champion per-facility distribution (bootstrap).

**Assumption:** champion and non-champion facilities have the same case load
per facility on average.

```{r method-a}
res_a <- burden_a(
  data            = d,
  target_elements = target_elements,
  region_col      = "Province",
  n_bootstrap     = 1000L
)
```

### Method C1 — Linear imputation

For each facility, fits a linear regression `value ~ champion_mean` on the
full available history. Missing months within the target period are imputed
from the model. Facilities too sparse to fit a model (< 3 observed months)
are handled via the champion-distribution fallback.

**Advantage over A:** uses facility-specific trends; rewards facilities with
partial but consistent reporting.

```{r method-c1}
res_c1 <- burden_c1(
  data            = d,
  target_elements = target_elements,
  region_col      = "Province",
  period_start    = period_start,
  period_end      = period_end,
  min_obs         = 3L,
  n_bootstrap     = 1000L
)
```

### Method C2 — ARIMA imputation

Replaces the linear model with `auto.arima`, using the regional champion mean
as an external regressor. Requires the `forecast` package. Falls back to C1
for facilities with fewer than `min_months` of history.

**Advantage over C1:** captures autocorrelation and seasonality; better for
facilities with long but irregular series.

```{r method-c2}
res_c2 <- burden_c2(
  data            = d,
  target_elements = target_elements,
  region_col      = "Province",
  period_start    = period_start,
  period_end      = period_end,
  min_months      = 12L,
  n_bootstrap     = 1000L
)
```

---

## 5  View results

### National-level summary table

```{r summary-national}
burden_summary_table(
  results_list = list(A = res_a, C1 = res_c1, C2 = res_c2),
  level        = "national"
)
```

```
    region                                 category                      A                     C1                    C2
1 National  MAL - Confirmed malaria cases_0-4 years   7,682 [5,405–11,567]   5,639 [4,231–7,525]  5,482 [4,189–7,244]
2 National  MAL - Confirmed malaria cases_15+ years 18,718 [13,407–29,555]  8,777 [7,844–10,488]  8,622 [7,744–9,812]
3 National MAL - Confirmed malaria cases_5-14 years   9,388 [7,616–12,493]   5,372 [4,884–6,139]  5,277 [4,812–5,965]
```

Estimates are shown as `point estimate [95% bootstrap CI]`.

### Province-level (subnational) table

```{r summary-subnational}
sub <- burden_summary_table(
  results_list = list(A = res_a, C1 = res_c1),
  level        = "subnational"
)
head(sub, 10)
```

```
                region                                 category                A                  C1
  01 Vientiane Capital  MAL - Confirmed malaria cases_0-4 years      90 [24–150]       102 [91–112]
  01 Vientiane Capital  MAL - Confirmed malaria cases_15+ years     120 [28–210]      134 [119–149]
  01 Vientiane Capital MAL - Confirmed malaria cases_5-14 years       56 [20–90]      124 [116–131]
          02 Phongsali  MAL - Confirmed malaria cases_0-4 years    653 [475–871]      234 [206–276]
          02 Phongsali  MAL - Confirmed malaria cases_15+ years   932 [497–1,751]     179 [179–179]
          02 Phongsali MAL - Confirmed malaria cases_5-14 years    565 [329–857]      167 [147–198]
       03 Louangnamtha  MAL - Confirmed malaria cases_0-4 years            <NA>       155 [15–573]
       03 Louangnamtha  MAL - Confirmed malaria cases_15+ years            <NA>       317 [102–998]
       03 Louangnamtha MAL - Confirmed malaria cases_5-14 years            <NA>       224 [31–751]
```

Provinces with `<NA>` under Method A had no champion facilities — C1 still
produces an estimate by predicting from the national champion distribution.

### Age-group totals

`add_category_totals()` sums across categories, propagating uncertainty
correctly (√ΣSE²):

```{r totals}
res_a_with_totals  <- add_category_totals(res_a)
res_c1_with_totals <- add_category_totals(res_c1)

burden_summary_table(
  list(A = res_a_with_totals, C1 = res_c1_with_totals),
  level = "national"
)
```

```
    region                                 category                      A                    C1
1 National  MAL - Confirmed malaria cases_0-4 years   7,682 [5,405–11,567]  5,639 [4,231–7,525]
2 National  MAL - Confirmed malaria cases_15+ years 18,718 [13,407–29,555] 8,777 [7,844–10,488]
3 National MAL - Confirmed malaria cases_5-14 years   9,388 [7,616–12,493]  5,372 [4,884–6,139]
4 National                                    Total 35,788 [26,428–53,615] 19,788 [16,959–24,152]
```

---

## 6  Interpreting the results

**Why do A and C1 differ?**
Method A assumes non-reporting facilities have the same per-facility rate as
champions. C1 uses each facility's own reporting history to impute missing months,
so facilities with genuinely low or zero case counts pull the estimate down.
In settings where zero is often stored as missing (e.g., a facility that sees no
malaria but submits data for other elements), C1 is more conservative and likely
closer to the truth.

**Provinces with no champions (e.g., 03 Louangnamtha)**
Method A returns `NA` — it cannot extrapolate without a regional reference.
C1 falls back to the national champion distribution as the prior for non-reporting
facilities, giving a wide interval that reflects genuine uncertainty.

**Reporting vs. estimated**
The reported total (sum of all submitted values from any facility) gives the
lower bound on actual burden. Estimates above this figure represent the
contribution of non- or inconsistently-reporting facilities.

---

## 7  Running all methods via the wrapper

`run_burden_estimates()` runs whichever subset of methods you specify and
returns a named list suitable for `burden_summary_table()`:

```{r wrapper}
results <- run_burden_estimates(
  data            = d,
  target_elements = target_elements,
  region_col      = "Province",
  methods         = c("A", "C1", "C2"),
  period_start    = period_start,
  period_end      = period_end,
  n_bootstrap     = 1000L
)

burden_summary_table(results, level = "national")
```

---

## Session info

```{r session}
sessionInfo()
```
