---
title: "Burden Estimation: Standalone Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Burden Estimation: Standalone Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)

# Shortens long DHIS2 element names for table display.
# "MAL - Confirmed malaria cases_0-4 years" → "Confirmed_0-4y"
fmt_cat <- function(x) {
  x <- sub("^[A-Z]+ - ", "", x)          # drop "MAL - " prefix
  x <- sub(" years?$", "y", x)           # "years" → "y"
  x <- sub(" (\\d)", "_\\1", x)          # space before age → underscore
  x
}

# Wrapper: run burden_summary_table and format for kable output
burden_kable <- function(results_list, level = "national", caption = NULL) {
  tbl <- burden_summary_table(results_list, level = level)
  if (is.null(tbl)) return(invisible(NULL))
  if ("category" %in% names(tbl)) tbl$category <- fmt_cat(tbl$category)
  knitr::kable(tbl, caption = caption, align = "llrrrr",
               format.args = list(big.mark = ","))
}
```

## Overview

Routine health facility reporting rarely captures all cases. Facilities may not
report every month, and some confirmed cases may be recorded as zero rather than
missing. The MG2 burden estimation module adjusts for these gaps by:

1. Identifying **champion facilities** — those that report consistently and are
   used as the reference population.
2. **Extrapolating** from champion rates to all facilities in each region using
   one of several statistical methods.
3. Producing **confidence intervals** via bootstrap resampling (1,000 draws).

This vignette uses the bundled demo dataset (Lao PDR malaria surveillance,
Aug 2021 – May 2026, from the public DHIS2 demo server at
<https://demos.dhis2.org>) to walk through the full standalone R workflow.

---

## 1  Locate the demo data

`mg2_demo_dir()` returns the path to the three bundled demo files:

```{r demo-dir}
library(MG2)
library(tsibble)     # yearmonth()
library(data.table)  # setDT(), :=

demo  <- mg2_demo_dir()
list.files(demo)
```

The directory mirrors a real MG2 data folder and can also be pointed at
directly from the Shiny app's Setup tab.

---

## 2  Load data, formula, and metadata

```{r load-all}
# Processed 6-year dataset (18 columns, ~867 k rows)
d_raw <- read_file(file.path(demo, "malaria_demo.rds"))

# Formula: maps data element IDs to human-readable names and categories
formula_file <- file.path(demo, "Formulas_demo.xlsx")
formula_elements <- readxl::read_excel(formula_file, sheet = "Formula Elements")

# Metadata: org unit tree, geographic features, validation rules, etc.
meta <- readRDS(file.path(demo, "metadata_demo.rds"))

cat("Data:    ", nrow(d_raw), "rows,", format(min(d_raw$Month)), "-",
    format(max(d_raw$Month)), "\n")
cat("Formula: ", nrow(formula_elements), "data elements\n")
cat("Metadata:", length(meta), "objects —", paste(names(meta), collapse = ", "), "\n")
cat("Provinces:", paste(sort(unique(d_raw$Province)), collapse = ", "), "\n")
```

---

## 3  Prepare data for analysis

`cleanedData()` filters to facility-level rows and creates the `dataCol`
numeric column from `original`.  `selectedData()` filters to the chosen
elements, identifies champion facilities, and adds a `Selected` column.

```{r prepare}
target_elements <- c(
  "MAL - Confirmed malaria cases_0-4 years",
  "MAL - Confirmed malaria cases_5-14 years",
  "MAL - Confirmed malaria cases_15+ years"
)

period_start <- yearmonth("2024 Jan")
period_end   <- yearmonth("2025 Dec")

# Keep only facility-level rows (level 4).
# cleanedData() retains all effectiveLeaf == TRUE rows — the correct set of
# data entry points regardless of administrative level.  District hospitals
# (level 3) that enter data directly are legitimate reporting units and must
# not be excluded by a level-4 filter.
d_clean <- cleanedData(d_raw, source = "Original")

# Champion identification uses malaria-specific elements (confirmed cases +
# diagnostic elements) so champion status reflects malaria-programme
# engagement.  Using all 67 elements would make nearly every facility a
# champion via outpatient attendance, collapsing the non-champion
# extrapolation and deflating estimates.
champ_elements <- grep(
  "Confirmed malaria|suspects tested|Suspected malaria|All-cause outpatients",
  unique(d_clean$data), ignore.case = TRUE, value = TRUE
)
# Outpatients are included so that Method B (attendance-based) has
# attendance data in d.  They also provide a reasonable basis for
# identifying active reporting facilities for champion selection.
cat("Champion identification elements:", length(champ_elements), "\n")

d <- selectedData(
  d_clean,
  levelNames      = c("Country", "Province", "District", "Facility"),
  data_categories = champ_elements,
  startingMonth   = period_start,
  endingMonth     = period_end,
  missing_reports = 2L
)

setDT(d)
d[, value := dataCol]

cat("Rows in d:", nrow(d), "\n")
cat("Unique facilities:", uniqueN(d$orgUnit), "\n")

# Champion summary
champ <- unique(d[Selected == "Champion", .(orgUnit, Province)])
cat("\nChampions by province:\n")
print(sort(table(champ$Province)))
```

---

## 4  Run burden estimates

### Method A — Champion multiple

Each non-champion facility is independently predicted from the champion
per-facility distribution in its region. Regions with no local champions
fall back to neighbouring provinces, then to the national distribution.

```{r method-a}
res_a <- burden_a(
  data            = d,
  target_elements = target_elements,
  region_col      = "Province",
  period_start    = period_start,
  period_end      = period_end,
  n_bootstrap     = 1000L
  # neighbor_list computed automatically from geo_features() inside the app;
  # pass explicitly here if you have an sf province polygon layer.
)
```

### Method C1 — Linear imputation (log scale)

For each facility, fits `log(1 + y) ~ champion_mean` on the full history and
imputes missing months in the estimate period.  Predictions are
back-transformed with `expm1()` so confidence intervals are always ≥ 0.

```{r method-c1}
res_c1 <- burden_c1(
  data            = d,
  target_elements = target_elements,
  region_col      = "Province",
  period_start    = period_start,
  period_end      = period_end,
  min_obs         = 3L,
  n_bootstrap     = 1000L
)
```

### Method C2 — ARIMA imputation (log scale)

Same as C1 but replaces the linear model with `auto.arima`. Requires the
`forecast` package; falls back to C1 for short series (< 12 months).

```{r method-c2}
res_c2 <- burden_c2(
  data            = d,
  target_elements = target_elements,
  region_col      = "Province",
  period_start    = period_start,
  period_end      = period_end,
  min_months      = 12L,
  n_bootstrap     = 1000L
)
```

### Method B — Attendance-based champion multiple

Method B replaces the per-facility resampling of Method A with a log-log
regression on attendance.  For each champion facility, fit
`log(cases) ~ log(attendance)`.  Then predict each non-champion facility's
expected case count from its own attendance data.

The age categories of all-cause outpatients match confirmed malaria cases
exactly (0–4 y, 5–14 y, 15+ y), so no category aggregation is needed.

```{r method-b}
# The DHIS2 formula often disaggregates attendance by age × sex, giving
# six elements (0-4 Female, 0-4 Male, 5-14 Female, …).  The demo data was
# downloaded with age-only totals already summed across sex, giving three
# elements.  The code below detects which form is present and pre-aggregates
# if sex-disaggregated data is found.

att_raw <- grep("All-cause outpatients", unique(d$data), value = TRUE)
att_has_sex <- any(grepl("Female|Male", att_raw))
cat("Attendance elements in data:", length(att_raw),
    if (att_has_sex) "(sex-disaggregated — will aggregate)" else "(age totals — ready to use)", "\n")

if (att_has_sex) {
  # Sum Male + Female for each age group and month, then bind back as
  # age-only rows so Method B can use a 1:1 category map.
  age_groups <- sub(", (Female|Male)$", "", unique(att_raw))
  age_groups <- unique(age_groups)   # e.g. "GEN - All-cause outpatients_0-4 years"

  # Identify all columns to group by (all except value / original)
  grp_cols <- setdiff(names(d), c("value", "original", "dataCol",
                                   "key_entry_error", "over_max",
                                   "mad15", "mad10", "mad5", "seasonal5", "seasonal3"))

  att_agg <- d[d$data %in% att_raw, ]
  att_agg$data <- sub(", (Female|Male)$", "", att_agg$data)
  att_agg <- aggregate(
    cbind(value = att_agg$value, original = att_agg$original),
    by  = att_agg[, intersect(grp_cols, names(att_agg)), drop = FALSE],
    FUN = function(x) sum(x, na.rm = TRUE)
  )
  # Add zero-filled outlier flag columns to match d's structure
  for (col in c("key_entry_error","over_max","mad15","mad10","mad5","seasonal5","seasonal3")) {
    if (col %in% names(d)) att_agg[[col]] <- FALSE
  }
  att_agg$dataCol <- att_agg$value
  d_b <- rbind(d[!d$data %in% att_raw, ], att_agg)
  attendance_elements <- age_groups
} else {
  d_b <- d
  attendance_elements <- att_raw
}

# 1:1 category map: each confirmed-case age group → its attendance counterpart
cat_map_b <- setNames(attendance_elements, target_elements)
cat("Category map:\n"); print(cat_map_b)

res_b <- burden_b(
  data                = d_b,
  target_elements     = target_elements,
  attendance_elements = attendance_elements,
  cat_map             = cat_map_b,
  region_col          = "Province",
  n_bootstrap         = 1000L
)
```

### Method E — Adjusted Corrected Incidence (Robust Estimator)

Method E (Thwing, Plucinski, Painter et al. 2020, *AJTMH* 102:811–820)
corrects confirmed case counts for patients who attended but were not tested
for malaria.  It requires three elements per facility: confirmed cases,
total attendance, and patients tested.

Applied **only to champion facilities** that consistently reported all three
elements.  The correction uses five parameters: β (proportion of non-malaria
fever consulters), α (ratio of TPR in untested vs. tested), γ (background
fever incidence), λ_min (minimum malaria-attributable fraction), and the
empirical testing rate T/A.  Default values follow Plucinski et al. 2018.

*Note on the tested element*: the RDT and microscopy suspects-tested elements
are sex-stratified (six sub-categories).  `MAL - Suspected malaria cases` —
patients presenting with fever and suspected of malaria — serves as the
testing denominator here.  In a full analysis you would sum
`suspects tested (RDT)` + `suspects tested (microscopy)` across sexes.

```{r method-e}
tested_elements <- c(
  "MAL - Suspected malaria cases_0-4 years",
  "MAL - Suspected malaria cases_5-14 years",
  "MAL - Suspected malaria cases_15+ years"
)

cat_map_attend <- setNames(attendance_elements, target_elements)
cat_map_tested <- setNames(tested_elements,    target_elements)

res_e <- burden_aci(
  data                = d_b,   # uses same age-aggregated attendance as Method B
  confirmed_elements  = target_elements,
  attendance_elements = attendance_elements,
  tested_elements     = tested_elements,
  cat_map_attend      = cat_map_attend,
  cat_map_tested      = cat_map_tested,
  region_col          = "Province",
  n_bootstrap         = 1000L
)
```

---

## 5  How C1 and C2 work — and why they differ from A

### C1 step-by-step

For each target element and each region, C1 proceeds as follows.

**Setup**

1. Compute the champion monthly mean: average value for each calendar month
   (Jan–Dec) across all champion facilities in the region.  The national mean
   is used as a fallback for months or regions with no local champion data.

**Main facility loop** — all facilities that submitted at least one
target-element record during the estimate period (Jan 2024–Dec 2025)

2. Retrieve the facility's *full* available history (all months in `d`, not
   just the period) to maximise model fitting data.
3. Fit `log(1 + value) ~ champion_monthly_mean` by ordinary least squares on
   all observed months.  Fitting and predicting on the log(1+x) scale ensures
   all imputed values are non-negative.
4. **If fewer than 3 observed months**: regression is skipped; the facility's
   actual submitted values are credited but missing months are not imputed
   (conservative — treated as 0).
5. For each month in the estimate period:
   - *Value present*: add the actual submitted value.
   - *Value missing*: draw from
     `max(0, exp(a + b × champion_mean + ε) − 1)`, where
     ε ~ Normal(0, σ²) and σ is the regression residual SD.  Repeated
     1,000 times to build the bootstrap distribution.
6. Facility contribution = sum across the 24 estimate-period months.

**No-data fallback** — facilities with zero target-element records at all

7. *Champion* facilities contribute **0** (they would have reported positives
   if any existed — their absence of confirmed cases is a true zero).
8. *Non-champion* facilities each independently draw one value from the
   champion per-period distribution (same mechanism as Method A).

**Region total** = sum over all facilities; repeat 1,000× for the CI.

### C2

Identical to C1 except step 3: when ≥ 12 months of history exist,
`auto.arima` on `log(1 + value)` with the champion mean as an external
regressor replaces the linear model, capturing autocorrelation and
seasonality.  Falls back to C1 when the series is too short or the ARIMA
fit fails.

### Why C1 ≤ A

Method A applies the same rule to every non-champion: predict its period
total by sampling from the champion per-facility distribution, regardless of
what the facility itself has historically reported.  Method C1 uses each
facility's own record as a starting point.  Non-champion facilities —
by definition less consistent reporters — typically have sparser confirmed-case
histories.  The linear model fitted on their data yields a lower predicted
rate than the champion mean, producing a more conservative estimate.

If non-reporters are genuinely lower-burden (e.g. facilities in
low-transmission catchments that simply miss months occasionally), C1 is
closer to the truth.  If they have similar burden but poor data systems, A
is more appropriate.  The two methods bracket the plausible range.

### The sparse-data caveat

Facilities with only 1–2 observed months in the full history cannot be
modelled (step 4 above).  Their actual values are included but their
unobserved months contribute 0.  Method A would predict the full champion
monthly rate for those same months.  This is the main driver of A > C1 when
many facilities have sparse target-element records.

### Method B — how it differs

Method B replaces the empirical champion-distribution resample with a
regression covariate: each non-champion facility's own attendance record.
The log-log regression on champions gives the relationship between attendance
and cases.  Non-champions with high attendance get higher predicted cases than
those with low attendance, which is more informative than assuming every
non-champion behaves like a randomly selected champion.  Non-champions with
no attendance data fall back to Method A resampling.

B tends to be between A and C1: more targeted than A (uses attendance as a
signal) but still extrapolates to every non-champion, whereas C1 uses each
facility's own case history.

### Method E — interpretation

Method E is not directly comparable to A, B, C1, or C2 as a total-burden
estimate.  It applies only to champion facilities and asks: given what
champions reported, how many cases were missed because febrile patients
attended but were not tested?  The result is the champion-sector corrected
total.  To convert to a full population estimate, multiply Method E by the
Method A or C1 non-champion scaling factor, or compare it alongside Methods
A–C to see how much of the gap between reported and estimated is attributable
to incomplete testing versus non-reporting facilities.

---

## 6  View results

### National summary

```{r summary-national}
# Champion summary
fac_univ_nat <- unique(as.data.frame(d)[, c("orgUnit", "Selected")])
n_champ_nat  <- sum(fac_univ_nat$Selected == "Champion")
n_total_nat  <- nrow(fac_univ_nat)
cat(sprintf("Champion facilities: %d of %d (%.0f%%)\n",
            n_champ_nat, n_total_nat, 100 * n_champ_nat / n_total_nat))

# Reported totals by category (actual submitted values, period only)
rep_nat <- d[d$data %in% target_elements &
             d$Month >= period_start & d$Month <= period_end &
             !is.na(d$original), ]
rep_by_cat <- tapply(rep_nat$original, rep_nat$data, sum)
rep_total  <- sum(rep_by_cat)
cat(sprintf("Reported (all facilities, target elements): %s cases\n",
            formatC(rep_total, format = "d", big.mark = ",")))

# Build national table with Reported column prepended
nat_tbl <- burden_summary_table(
  lapply(list(A = res_a, B = res_b, C1 = res_c1, C2 = res_c2, E = res_e),
         add_category_totals),
  level = "national"
)
nat_tbl$category <- fmt_cat(nat_tbl$category)

# Add Reported column
rep_df <- data.frame(
  category = fmt_cat(names(rep_by_cat)),
  Reported = formatC(as.integer(rep_by_cat), format = "d", big.mark = ","),
  stringsAsFactors = FALSE
)
rep_total_row <- data.frame(category = "Total",
                             Reported = formatC(rep_total, format="d", big.mark=","),
                             stringsAsFactors = FALSE)
rep_df <- rbind(rep_df, rep_total_row)
nat_tbl <- merge(rep_df, nat_tbl[, setdiff(names(nat_tbl), "region")],
                 by = "category", all.y = TRUE)

knitr::kable(nat_tbl,
  caption = sprintf(
    "National burden estimates — Jan 2024 to Dec 2025 (%d/%d champion facilities)",
    n_champ_nat, n_total_nat))
```

### Province-level table

```{r summary-subnational}
# Champion / total facility counts — one row per unique facility
fac_univ <- unique(as.data.frame(d)[, c("orgUnit", "Province", "Selected")])
champ_tbl <- do.call(rbind, lapply(split(fac_univ, fac_univ$Province), function(x) {
  data.frame(
    region       = x$Province[1],
    `Champ/Total` = paste0(sum(x$Selected == "Champion"), "/", nrow(x)),
    check.names = FALSE
  )
}))

# National champion count
n_champ <- sum(fac_univ$Selected == "Champion")
n_total <- nrow(fac_univ)
nat_champ <- data.frame(region = "National",
                        `Champ/Total` = paste0(n_champ, "/", n_total),
                        check.names = FALSE)

# Build subnational table — keep only Total row per province
tbl_sub <- burden_summary_table(
  lapply(list(A = res_a, B = res_b, C1 = res_c1, C2 = res_c2, E = res_e),
         add_category_totals),
  level = "subnational"
)
tbl_sub <- tbl_sub[tbl_sub$category == "Total", ]
tbl_sub$category <- NULL

# Merge in champion counts
tbl_sub <- merge(champ_tbl, tbl_sub, by = "region", all.y = TRUE)
knitr::kable(tbl_sub,
             caption = "Province-level totals — all ages combined",
             col.names = c("Province", "Champ/Total", "A", "B", "C1", "C2", "E"))
```

---

## 6  Choropleth maps

Three side-by-side province maps compare what was actually reported against
what Method A estimates, and then split provinces into two significantly
different burden groups.

```{r choropleths, fig.width = 9, fig.height = 6.5, out.width = "100%", fig.align = "center"}
library(ggplot2)
library(sf)
library(patchwork)

# Province polygons from metadata
provinces <- meta$geoFeatures[
  meta$geoFeatures$level == 2 & !sf::st_is_empty(meta$geoFeatures), ]

# Reported: sum of target-element values only (Jan 2024–Dec 2025)
rep_by_prov <- d[data %in% target_elements &
                 Month >= period_start & Month <= period_end & !is.na(original),
                 .(reported = sum(original, na.rm = TRUE)), by = Province]

# Method A and C2 totals: sum across age groups per province
sum_sub <- function(res, col_name) {
  df <- as.data.frame(res$subnational)
  df <- df[df$region != "National", ]
  df <- aggregate(estimate ~ region, data = df, sum)
  names(df) <- c("name", col_name)
  df
}
a_total  <- sum_sub(res_a,  "estimate_a")
c2_total <- sum_sub(res_c2, "estimate_c2")

# Significance groups from Method A (2 groups, ward.D2 on Z-score distance)
grps   <- compute_burden_groups(res_a$subnational, n_groups = 2)
grp_df <- data.frame(name  = names(grps),
                     group = factor(grps, labels = c("Lower", "Higher")))

# Join all to sf
map_sf <- merge(provinces,   rep_by_prov, by.x = "name", by.y = "Province", all.x = TRUE)
map_sf <- merge(map_sf, a_total,          by = "name", all.x = TRUE)
map_sf <- merge(map_sf, c2_total,         by = "name", all.x = TRUE)
map_sf <- merge(map_sf, grp_df,           by = "name", all.x = TRUE)

# Shared upper limit for the three continuous maps — same scale makes the
# gap between reported and modelled visually obvious.
shared_max    <- max(c(map_sf$reported, map_sf$estimate_a, map_sf$estimate_c2),
                     na.rm = TRUE)
shared_limits <- c(0, shared_max)

theme_map <- theme_void(base_size = 9) +
  theme(plot.title      = element_text(hjust = 0.5, face = "bold", size = 10),
        legend.position = "bottom",
        legend.key.width = unit(1.8, "cm"),
        legend.key.height = unit(0.35, "cm"),
        legend.title    = element_text(size = 8),
        legend.text     = element_text(size = 7))

# Shared continuous scale — identical across p1/p2/p3 so patchwork
# merges them into a single legend when guides = "collect"
fill_shared <- function() {
  scale_fill_gradientn(
    colours  = c("#fff5eb", "#fd8d3c", "#8c2d04"),
    limits   = shared_limits,
    oob      = scales::squish,
    na.value = "#dddddd",
    name     = "Cases (all ages)",
    labels   = scales::label_comma(),
    guide    = guide_colorbar(title.position = "top",
                              barwidth = 10, barheight = 0.5)
  )
}

p1 <- ggplot(map_sf) +
  geom_sf(aes(fill = reported),    colour = "white", linewidth = 0.25) +
  fill_shared() + labs(title = "Reported") + theme_map

p2 <- ggplot(map_sf) +
  geom_sf(aes(fill = estimate_a),  colour = "white", linewidth = 0.25) +
  fill_shared() + labs(title = "Method A estimate") + theme_map

p3 <- ggplot(map_sf) +
  geom_sf(aes(fill = estimate_c2), colour = "white", linewidth = 0.25) +
  fill_shared() + labs(title = "Method C2 estimate (ARIMA)") + theme_map

p4 <- ggplot(map_sf) +
  geom_sf(aes(fill = group), colour = "white", linewidth = 0.25) +
  scale_fill_manual(
    values   = c("Lower" = "#fdd0a2", "Higher" = "#d94801"),
    na.value = "#dddddd",
    name     = "Burden group",
    guide    = guide_legend(title.position = "top")
  ) +
  labs(title = "Method A groups (k = 2)") + theme_map

# 2×2 grid; shared continuous legend appears once at bottom
(p1 + p2) / (p3 + p4) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

---

## 7  Run all methods via the wrapper

`run_burden_estimates()` runs whichever subset of methods you specify and
returns a named list ready for `burden_summary_table()`.

```{r wrapper}
# run_burden_estimates() uses the full d; period filtering happens inside
# each method.  Pass the same d used above (already filtered to the period
# by selectedData()).
results <- run_burden_estimates(
  data                = d_b,
  target_elements     = target_elements,
  attendance_elements = attendance_elements,
  tested_elements     = tested_elements,
  cat_map_attend      = cat_map_attend,
  cat_map_tested      = cat_map_tested,
  region_col          = "Province",
  methods             = c("A", "B", "C1", "C2", "E"),
  n_bootstrap         = 1000L
)

method_results <- results[intersect(c("A", "B", "C1", "C2", "E"), names(results))]
burden_kable(lapply(method_results, add_category_totals),
             level = "national",
             caption = "Wrapper — all five methods")
```

---

## Session info

```{r session}
sessionInfo()
```
