---
title: "Burden Estimation: Standalone Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Burden Estimation: Standalone Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)

# Shortens long DHIS2 element names for table display.
# "MAL - Confirmed malaria cases_0-4 years" → "Confirmed_0-4y"
fmt_cat <- function(x) {
  x <- sub("^[A-Z]+ - ", "", x)          # drop "MAL - " prefix
  x <- sub(" years?$", "y", x)           # "years" → "y"
  x <- sub(" (\\d)", "_\\1", x)          # space before age → underscore
  x
}

# Wrapper: run burden_summary_table and format for kable output
burden_kable <- function(results_list, level = "national", caption = NULL) {
  tbl <- burden_summary_table(results_list, level = level)
  if (is.null(tbl)) return(invisible(NULL))
  if ("category" %in% names(tbl)) tbl$category <- fmt_cat(tbl$category)
  knitr::kable(tbl, caption = caption, align = "llrrrr",
               format.args = list(big.mark = ","))
}
```

## Overview

Routine health facility reporting rarely captures all cases. Facilities may not
report every month, and some confirmed cases may be recorded as zero rather than
missing. The MG2 burden estimation module adjusts for these gaps by:

1. Identifying **champion facilities** — those that report consistently and are
   used as the reference population.
2. **Extrapolating** from champion rates to all facilities in each region using
   one of several statistical methods.
3. Producing **confidence intervals** via bootstrap resampling (1,000 draws).

This vignette uses the bundled demo dataset (Lao PDR malaria surveillance,
Aug 2021 – May 2026, from the public DHIS2 demo server at
<https://demos.dhis2.org>) to walk through the full standalone R workflow.

---

## 1  Locate the demo data

`mg2_demo_dir()` returns the path to the three bundled demo files:

```{r demo-dir}
library(MG2)
library(tsibble)     # yearmonth()
library(data.table)  # setDT(), :=

demo  <- mg2_demo_dir()
list.files(demo)
```

The directory mirrors a real MG2 data folder and can also be pointed at
directly from the Shiny app's Setup tab.

---

## 2  Load data, formula, and metadata

```{r load-all}
# Processed 6-year dataset (18 columns, ~867 k rows)
d_raw <- read_file(file.path(demo, "malaria_demo.rds"))

# Formula: maps data element IDs to human-readable names and categories
formula_file <- file.path(demo, "Formulas_demo.xlsx")
formula_elements <- readxl::read_excel(formula_file, sheet = "Formula Elements")

# Metadata: org unit tree, geographic features, validation rules, etc.
meta <- readRDS(file.path(demo, "metadata_demo.rds"))

cat("Data:    ", nrow(d_raw), "rows,", format(min(d_raw$Month)), "-",
    format(max(d_raw$Month)), "\n")
cat("Formula: ", nrow(formula_elements), "data elements\n")
cat("Metadata:", length(meta), "objects —", paste(names(meta), collapse = ", "), "\n")
cat("Provinces:", paste(sort(unique(d_raw$Province)), collapse = ", "), "\n")
```

---

## 3  Prepare data for analysis

`cleanedData()` filters to facility-level rows and creates the `dataCol`
numeric column from `original`.  `selectedData()` filters to the chosen
elements, identifies champion facilities, and adds a `Selected` column.

```{r prepare}
target_elements <- c(
  "MAL - Confirmed malaria cases_0-4 years",
  "MAL - Confirmed malaria cases_5-14 years",
  "MAL - Confirmed malaria cases_15+ years"
)

period_start <- yearmonth("2024 Jan")
period_end   <- yearmonth("2025 Dec")

# Keep only facility-level rows (level 4).
# cleanedData() retains all effectiveLeaf == TRUE rows — the correct set of
# data entry points regardless of administrative level.  District hospitals
# (level 3) that enter data directly are legitimate reporting units and must
# not be excluded by a level-4 filter.
d_clean <- cleanedData(d_raw, source = "Original")

# Champion identification uses malaria-specific elements (confirmed cases +
# diagnostic elements) so champion status reflects malaria-programme
# engagement.  Using all 67 elements would make nearly every facility a
# champion via outpatient attendance, collapsing the non-champion
# extrapolation and deflating estimates.
champ_elements <- grep(
  "Confirmed malaria|suspects tested|Suspected malaria",
  unique(d_clean$data), ignore.case = TRUE, value = TRUE
)
cat("Champion identification elements:", length(champ_elements), "\n")

d <- selectedData(
  d_clean,
  levelNames      = c("Country", "Province", "District", "Facility"),
  data_categories = champ_elements,
  startingMonth   = period_start,
  endingMonth     = period_end,
  missing_reports = 2L
)

setDT(d)
d[, value := dataCol]

cat("Rows in d:", nrow(d), "\n")
cat("Unique facilities:", uniqueN(d$orgUnit), "\n")

# Champion summary
champ <- unique(d[Selected == "Champion", .(orgUnit, Province)])
cat("\nChampions by province:\n")
print(sort(table(champ$Province)))
```

---

## 4  Run burden estimates

### Method A — Champion multiple

Each non-champion facility is independently predicted from the champion
per-facility distribution in its region. Regions with no local champions
fall back to neighbouring provinces, then to the national distribution.

```{r method-a}
res_a <- burden_a(
  data            = d,
  target_elements = target_elements,
  region_col      = "Province",
  period_start    = period_start,
  period_end      = period_end,
  n_bootstrap     = 1000L
  # neighbor_list computed automatically from geo_features() inside the app;
  # pass explicitly here if you have an sf province polygon layer.
)
```

### Method C1 — Linear imputation (log scale)

For each facility, fits `log(1 + y) ~ champion_mean` on the full history and
imputes missing months in the estimate period.  Predictions are
back-transformed with `expm1()` so confidence intervals are always ≥ 0.

```{r method-c1}
res_c1 <- burden_c1(
  data            = d,
  target_elements = target_elements,
  region_col      = "Province",
  period_start    = period_start,
  period_end      = period_end,
  min_obs         = 3L,
  n_bootstrap     = 1000L
)
```

### Method C2 — ARIMA imputation (log scale)

Same as C1 but replaces the linear model with `auto.arima`. Requires the
`forecast` package; falls back to C1 for short series (< 12 months).

```{r method-c2}
res_c2 <- burden_c2(
  data            = d,
  target_elements = target_elements,
  region_col      = "Province",
  period_start    = period_start,
  period_end      = period_end,
  min_months      = 12L,
  n_bootstrap     = 1000L
)
```

---

## 5  View results

### National summary

```{r summary-national}
# Champion summary
fac_univ_nat <- unique(as.data.frame(d)[, c("orgUnit", "Selected")])
n_champ_nat  <- sum(fac_univ_nat$Selected == "Champion")
n_total_nat  <- nrow(fac_univ_nat)
cat(sprintf("Champion facilities: %d of %d (%.0f%%)\n",
            n_champ_nat, n_total_nat, 100 * n_champ_nat / n_total_nat))

# Reported totals by category (actual submitted values, period only)
rep_nat <- d[d$data %in% target_elements &
             d$Month >= period_start & d$Month <= period_end &
             !is.na(d$original), ]
rep_by_cat <- tapply(rep_nat$original, rep_nat$data, sum)
rep_total  <- sum(rep_by_cat)
cat(sprintf("Reported (all facilities, target elements): %s cases\n",
            formatC(rep_total, format = "d", big.mark = ",")))

# Build national table with Reported column prepended
nat_tbl <- burden_summary_table(
  lapply(list(A = res_a, C1 = res_c1, C2 = res_c2), add_category_totals),
  level = "national"
)
nat_tbl$category <- fmt_cat(nat_tbl$category)

# Add Reported column
rep_df <- data.frame(
  category = fmt_cat(names(rep_by_cat)),
  Reported = formatC(as.integer(rep_by_cat), format = "d", big.mark = ","),
  stringsAsFactors = FALSE
)
rep_total_row <- data.frame(category = "Total",
                             Reported = formatC(rep_total, format="d", big.mark=","),
                             stringsAsFactors = FALSE)
rep_df <- rbind(rep_df, rep_total_row)
nat_tbl <- merge(rep_df, nat_tbl[, setdiff(names(nat_tbl), "region")],
                 by = "category", all.y = TRUE)

knitr::kable(nat_tbl,
  caption = sprintf(
    "National burden estimates — Jan 2024 to Dec 2025 (%d/%d champion facilities)",
    n_champ_nat, n_total_nat))
```

### Province-level table

```{r summary-subnational}
# Champion / total facility counts — one row per unique facility
fac_univ <- unique(as.data.frame(d)[, c("orgUnit", "Province", "Selected")])
champ_tbl <- do.call(rbind, lapply(split(fac_univ, fac_univ$Province), function(x) {
  data.frame(
    region       = x$Province[1],
    `Champ/Total` = paste0(sum(x$Selected == "Champion"), "/", nrow(x)),
    check.names = FALSE
  )
}))

# National champion count
n_champ <- sum(fac_univ$Selected == "Champion")
n_total <- nrow(fac_univ)
nat_champ <- data.frame(region = "National",
                        `Champ/Total` = paste0(n_champ, "/", n_total),
                        check.names = FALSE)

# Build subnational table — keep only Total row per province
tbl_sub <- burden_summary_table(
  lapply(list(A = res_a, C1 = res_c1, C2 = res_c2), add_category_totals),
  level = "subnational"
)
tbl_sub <- tbl_sub[tbl_sub$category == "Total", ]
tbl_sub$category <- NULL

# Merge in champion counts
tbl_sub <- merge(champ_tbl, tbl_sub, by = "region", all.y = TRUE)
knitr::kable(tbl_sub,
             caption = "Province-level totals — all ages combined",
             col.names = c("Province", "Champ/Total", "A", "C1", "C2"))
```

---

## 6  Choropleth maps

Three side-by-side province maps compare what was actually reported against
what Method A estimates, and then split provinces into two significantly
different burden groups.

```{r choropleths, fig.width = 9, fig.height = 6.5, out.width = "100%", fig.align = "center"}
library(ggplot2)
library(sf)
library(patchwork)

# Province polygons from metadata
provinces <- meta$geoFeatures[
  meta$geoFeatures$level == 2 & !sf::st_is_empty(meta$geoFeatures), ]

# Reported: sum of target-element values only (Jan 2024–Dec 2025)
rep_by_prov <- d[data %in% target_elements &
                 Month >= period_start & Month <= period_end & !is.na(original),
                 .(reported = sum(original, na.rm = TRUE)), by = Province]

# Method A and C2 totals: sum across age groups per province
sum_sub <- function(res, col_name) {
  df <- as.data.frame(res$subnational)
  df <- df[df$region != "National", ]
  df <- aggregate(estimate ~ region, data = df, sum)
  names(df) <- c("name", col_name)
  df
}
a_total  <- sum_sub(res_a,  "estimate_a")
c2_total <- sum_sub(res_c2, "estimate_c2")

# Significance groups from Method A (2 groups, ward.D2 on Z-score distance)
grps   <- compute_burden_groups(res_a$subnational, n_groups = 2)
grp_df <- data.frame(name  = names(grps),
                     group = factor(grps, labels = c("Lower", "Higher")))

# Join all to sf
map_sf <- merge(provinces,   rep_by_prov, by.x = "name", by.y = "Province", all.x = TRUE)
map_sf <- merge(map_sf, a_total,          by = "name", all.x = TRUE)
map_sf <- merge(map_sf, c2_total,         by = "name", all.x = TRUE)
map_sf <- merge(map_sf, grp_df,           by = "name", all.x = TRUE)

# Shared upper limit for the three continuous maps — same scale makes the
# gap between reported and modelled visually obvious.
shared_max    <- max(c(map_sf$reported, map_sf$estimate_a, map_sf$estimate_c2),
                     na.rm = TRUE)
shared_limits <- c(0, shared_max)

theme_map <- theme_void(base_size = 9) +
  theme(plot.title      = element_text(hjust = 0.5, face = "bold", size = 10),
        legend.position = "bottom",
        legend.key.width = unit(1.8, "cm"),
        legend.key.height = unit(0.35, "cm"),
        legend.title    = element_text(size = 8),
        legend.text     = element_text(size = 7))

# Shared continuous scale — identical across p1/p2/p3 so patchwork
# merges them into a single legend when guides = "collect"
fill_shared <- function() {
  scale_fill_gradientn(
    colours  = c("#fff5eb", "#fd8d3c", "#8c2d04"),
    limits   = shared_limits,
    oob      = scales::squish,
    na.value = "#dddddd",
    name     = "Cases (all ages)",
    labels   = scales::label_comma(),
    guide    = guide_colorbar(title.position = "top",
                              barwidth = 10, barheight = 0.5)
  )
}

p1 <- ggplot(map_sf) +
  geom_sf(aes(fill = reported),    colour = "white", linewidth = 0.25) +
  fill_shared() + labs(title = "Reported") + theme_map

p2 <- ggplot(map_sf) +
  geom_sf(aes(fill = estimate_a),  colour = "white", linewidth = 0.25) +
  fill_shared() + labs(title = "Method A estimate") + theme_map

p3 <- ggplot(map_sf) +
  geom_sf(aes(fill = estimate_c2), colour = "white", linewidth = 0.25) +
  fill_shared() + labs(title = "Method C2 estimate (ARIMA)") + theme_map

p4 <- ggplot(map_sf) +
  geom_sf(aes(fill = group), colour = "white", linewidth = 0.25) +
  scale_fill_manual(
    values   = c("Lower" = "#fdd0a2", "Higher" = "#d94801"),
    na.value = "#dddddd",
    name     = "Burden group",
    guide    = guide_legend(title.position = "top")
  ) +
  labs(title = "Method A groups (k = 2)") + theme_map

# 2×2 grid; shared continuous legend appears once at bottom
(p1 + p2) / (p3 + p4) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

---

## 7  Run all methods via the wrapper

`run_burden_estimates()` runs whichever subset of methods you specify and
returns a named list ready for `burden_summary_table()`.

```{r wrapper}
# run_burden_estimates() uses the full d; period filtering happens inside
# each method.  Pass the same d used above (already filtered to the period
# by selectedData()).
results <- run_burden_estimates(
  data            = d,
  target_elements = target_elements,
  region_col      = "Province",
  methods         = c("A", "C1"),
  n_bootstrap     = 500L   # reduce for speed in this example
)

# run_burden_estimates() returns method results plus pre-built summary tables
# and a log vector.  Pass only the method entries to burden_summary_table().
method_results <- results[intersect(c("A", "B", "C1", "C2", "E"), names(results))]
burden_kable(method_results, level = "national",
             caption = "Wrapper output — Methods A and C1")
```

---

## Session info

```{r session}
sessionInfo()
```
