---
title: "Burden Estimation: Standalone Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Burden Estimation: Standalone Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)

# Shortens long DHIS2 element names for table display.
# "MAL - Confirmed malaria cases_0-4 years" → "Confirmed_0-4y"
fmt_cat <- function(x) {
  x <- sub("^[A-Z]+ - ", "", x)          # drop "MAL - " prefix
  x <- sub(" years?$", "y", x)           # "years" → "y"
  x <- sub(" (\\d)", "_\\1", x)          # space before age → underscore
  x
}

# Wrapper: run burden_summary_table and format for kable output
burden_kable <- function(results_list, level = "national", caption = NULL) {
  tbl <- burden_summary_table(results_list, level = level)
  if (is.null(tbl)) return(invisible(NULL))
  if ("category" %in% names(tbl)) tbl$category <- fmt_cat(tbl$category)
  knitr::kable(tbl, caption = caption, align = "llrrrr",
               format.args = list(big.mark = ","))
}
```

## Overview

Routine health facility reporting rarely captures all cases. Facilities may not
report every month, and some confirmed cases may be recorded as zero rather than
missing. The MG2 burden estimation module adjusts for these gaps by:

1. Identifying **champion facilities** — those that report consistently and are
   used as the reference population.
2. **Extrapolating** from champion rates to all facilities in each region using
   one of several statistical methods.
3. Producing **confidence intervals** via bootstrap resampling (1,000 draws).

This vignette uses the bundled demo dataset (Lao PDR malaria surveillance,
Aug 2021 – May 2026, from the public DHIS2 demo server at
<https://demos.dhis2.org>) to walk through the full standalone R workflow.

---

## 1  Locate the demo data

`mg2_demo_dir()` returns the path to the three bundled demo files:

```{r demo-dir}
library(MG2)
library(tsibble)     # yearmonth()
library(data.table)  # setDT(), :=

demo  <- mg2_demo_dir()
list.files(demo)
```

The directory mirrors a real MG2 data folder and can also be pointed at
directly from the Shiny app's Setup tab.

---

## 2  Load data, formula, and metadata

```{r load-all}
# Processed 6-year dataset (18 columns, ~867 k rows)
d_raw <- read_file(file.path(demo, "malaria_demo.rds"))

# Formula: maps data element IDs to human-readable names and categories
formula_file <- file.path(demo, "Formulas_demo.xlsx")
formula_elements <- readxl::read_excel(formula_file, sheet = "Formula Elements")

# Metadata: org unit tree, geographic features, validation rules, etc.
meta <- readRDS(file.path(demo, "metadata_demo.rds"))

cat("Data:    ", nrow(d_raw), "rows,", format(min(d_raw$Month)), "-",
    format(max(d_raw$Month)), "\n")
cat("Formula: ", nrow(formula_elements), "data elements\n")
cat("Metadata:", length(meta), "objects —", paste(names(meta), collapse = ", "), "\n")
cat("Provinces:", paste(sort(unique(d_raw$Province)), collapse = ", "), "\n")
```

---

## 3  Prepare data for analysis

`cleanedData()` filters to facility-level rows and creates the `dataCol`
numeric column from `original`.  `selectedData()` filters to the chosen
elements, identifies champion facilities, and adds a `Selected` column.

```{r prepare}
target_elements <- c(
  "MAL - Confirmed malaria cases_0-4 years",
  "MAL - Confirmed malaria cases_5-14 years",
  "MAL - Confirmed malaria cases_15+ years"
)

period_start <- yearmonth("2024 Jan")
period_end   <- yearmonth("2025 Dec")

d_clean <- cleanedData(d_raw, source = "Original")

d <- selectedData(
  d_clean,
  levelNames      = c("Country", "Province", "District", "Facility"),
  data_categories = target_elements,
  startingMonth   = period_start,
  endingMonth     = period_end,
  missing_reports = 2L   # allow up to 2 missed months in a 24-month window
)

# Rename dataCol → value for the burden functions
setDT(d)
d[, value := dataCol]

# Champion summary
champ <- unique(d[Selected == "Champion", .(orgUnit, Province)])
cat("Champions by province:\n")
print(sort(table(champ$Province)))
```

---

## 4  Run burden estimates

### Method A — Champion multiple

Each non-champion facility is independently predicted from the champion
per-facility distribution in its region. Regions with no local champions
fall back to neighbouring provinces, then to the national distribution.

```{r method-a}
res_a <- burden_a(
  data            = d,
  target_elements = target_elements,
  region_col      = "Province",
  period_start    = period_start,
  period_end      = period_end,
  n_bootstrap     = 1000L
  # neighbor_list computed automatically from geo_features() inside the app;
  # pass explicitly here if you have an sf province polygon layer.
)
```

### Method C1 — Linear imputation (log scale)

For each facility, fits `log(1 + y) ~ champion_mean` on the full history and
imputes missing months in the estimate period.  Predictions are
back-transformed with `expm1()` so confidence intervals are always ≥ 0.

```{r method-c1}
res_c1 <- burden_c1(
  data            = d,
  target_elements = target_elements,
  region_col      = "Province",
  period_start    = period_start,
  period_end      = period_end,
  min_obs         = 3L,
  n_bootstrap     = 1000L
)
```

### Method C2 — ARIMA imputation (log scale)

Same as C1 but replaces the linear model with `auto.arima`. Requires the
`forecast` package; falls back to C1 for short series (< 12 months).

```{r method-c2}
res_c2 <- burden_c2(
  data            = d,
  target_elements = target_elements,
  region_col      = "Province",
  period_start    = period_start,
  period_end      = period_end,
  min_months      = 12L,
  n_bootstrap     = 1000L
)
```

---

## 5  View results

### National summary

```{r summary-national}
burden_kable(
  lapply(list(A = res_a, C1 = res_c1, C2 = res_c2), add_category_totals),
  level   = "national",
  caption = "National burden estimates — Jan 2024 to Dec 2025"
)
```

### Province-level table

```{r summary-subnational}
# Build the subnational table with totals added, then keep only the
# "Total" row per province (sum across all age groups).
tbl_sub <- burden_summary_table(
  lapply(list(A = res_a, C1 = res_c1, C2 = res_c2), add_category_totals),
  level = "subnational"
)
tbl_sub <- tbl_sub[tbl_sub$category == "Total", ]
tbl_sub$category <- NULL   # redundant column now that every row is "Total"
knitr::kable(tbl_sub, caption = "Province-level totals — all ages combined")
```

### Age-group totals

`add_category_totals()` sums across categories, propagating uncertainty
as √ΣSE²:

```{r totals}
results_with_totals <- lapply(
  list(A = res_a, C1 = res_c1, C2 = res_c2),
  add_category_totals
)
burden_kable(results_with_totals, level = "national",
             caption = "National totals including all-age combined row")
```

---

## 6  Choropleth maps

Three side-by-side province maps compare what was actually reported against
what Method A estimates, and then split provinces into two significantly
different burden groups.

```{r choropleths, fig.width = 9, fig.height = 3.5, out.width = "100%"}
library(ggplot2)
library(sf)
library(patchwork)

# Province polygons from metadata
provinces <- meta$geoFeatures[
  meta$geoFeatures$level == 2 & !sf::st_is_empty(meta$geoFeatures), ]

# Reported: sum of all submitted values in the period
rep_by_prov <- d[!is.na(original),
                 .(reported = sum(original, na.rm = TRUE)), by = Province]

# Method A totals: sum across age groups
a_total <- as.data.frame(res_a$subnational)
a_total <- a_total[a_total$region != "National", ]
a_total <- aggregate(estimate ~ region, data = a_total, sum)
names(a_total) <- c("name", "estimate_a")

# Significance groups (2 groups via hierarchical clustering on Z-score distance)
grps    <- compute_burden_groups(res_a$subnational, n_groups = 2)
grp_df  <- data.frame(name  = names(grps),
                      group = factor(grps, labels = c("Lower", "Higher")))

# Join to sf
map_sf <- merge(provinces, rep_by_prov,
                by.x = "name", by.y = "Province", all.x = TRUE)
map_sf <- merge(map_sf, a_total,  by = "name", all.x = TRUE)
map_sf <- merge(map_sf, grp_df,   by = "name", all.x = TRUE)

# Shared helpers
theme_map <- theme_void(base_size = 9) +
  theme(plot.title    = element_text(hjust = 0.5, face = "bold"),
        legend.position = "bottom",
        legend.key.width = unit(1.2, "cm"))

fill_yor <- function(col, lab) {
  scale_fill_gradientn(
    aesthetics = "fill",
    colours    = c("#fff5eb", "#fd8d3c", "#8c2d04"),
    na.value   = "#dddddd",
    name       = lab,
    labels     = scales::label_comma(),
    guide      = guide_colorbar(title.position = "top")
  )
}

# Map 1: Reported cases
p1 <- ggplot(map_sf) +
  geom_sf(aes(fill = reported), colour = "white", linewidth = 0.25) +
  fill_yor("reported", "Cases") +
  labs(title = "Reported") + theme_map

# Map 2: Method A estimate
p2 <- ggplot(map_sf) +
  geom_sf(aes(fill = estimate_a), colour = "white", linewidth = 0.25) +
  fill_yor("estimate_a", "Cases") +
  labs(title = "Method A estimate") + theme_map

# Map 3: Two significance groups (ward.D2 on pairwise Z-score distance)
p3 <- ggplot(map_sf) +
  geom_sf(aes(fill = group), colour = "white", linewidth = 0.25) +
  scale_fill_manual(
    values   = c("Lower" = "#fdd0a2", "Higher" = "#d94801"),
    na.value = "#dddddd",
    name     = "Burden group",
    guide    = guide_legend(title.position = "top")
  ) +
  labs(title = "Groups (k = 2)") + theme_map

p1 + p2 + p3 + plot_annotation(
  caption = paste("Province-level estimates, Jan 2024–Dec 2025.",
                  "Groups derived from hierarchical clustering on pairwise",
                  "Z-score distance between Method A bootstrap distributions.")
)
```

---

## 7  Using the metadata

The metadata object provides geographic features for mapping and the org unit
hierarchy for filtering.

```{r metadata-use}
# Geographic features (sf): provinces, districts, facilities
gf <- meta$geoFeatures
cat("Geometry types:", paste(unique(sf::st_geometry_type(gf)), collapse = ", "), "\n")
cat("Levels:", paste(sort(unique(gf$level)), collapse = ", "), "\n")
cat(nrow(provinces), "province polygons (level 2)\n")

# Org unit tree: parent-child relationships
cat("\nOrg unit levels:\n")
knitr::kable(meta$orgUnitLevels[, c("level", "displayName")])
```

---

## 7  Run all methods via the wrapper

`run_burden_estimates()` runs whichever subset of methods you specify and
returns a named list ready for `burden_summary_table()`.

```{r wrapper}
# run_burden_estimates() uses the full d; period filtering happens inside
# each method.  Pass the same d used above (already filtered to the period
# by selectedData()).
results <- run_burden_estimates(
  data            = d,
  target_elements = target_elements,
  region_col      = "Province",
  methods         = c("A", "C1"),
  n_bootstrap     = 500L   # reduce for speed in this example
)

# run_burden_estimates() returns method results plus pre-built summary tables
# and a log vector.  Pass only the method entries to burden_summary_table().
method_results <- results[intersect(c("A", "B", "C1", "C2", "E"), names(results))]
burden_kable(method_results, level = "national",
             caption = "Wrapper output — Methods A and C1")
```

---

## Session info

```{r session}
sessionInfo()
```
