---
title: "Facility Counts and Data Element Reporting in DHIS2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Facility Counts and Data Element Reporting in DHIS2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## The question no one asks

When a DHIS2 report shows 500 confirmed malaria cases in a district last month,
an analyst naturally asks: *Is that more or fewer than last month?* But there is
a prior question that is almost never asked: *How many facilities contributed to
that 500?*

If 40 facilities reported in January and 25 reported in February, a drop in total
cases may reflect improved malaria control — or it may simply reflect 15 facilities
that did not submit data. Without knowing the number of contributing facilities,
the two explanations are indistinguishable.

This vignette explains why facility counts are harder to obtain than they appear
in DHIS2, what the MG2 API download actually returns, and how MG2 resolves the
counting ambiguity. The companion vignette *Champion Facilities and Reporting Bias*
uses these counts to identify consistently reporting facilities and correct for
reporting bias.

---

## 1  Two types of reporting completeness

DHIS2 tracks completeness at two levels, and they measure very different things.

**Dataset completeness** (form-level) counts whether the data entry form was
submitted. A facility is recorded as "complete" when the data entry clerk presses
the *Complete* button in DHIS2, regardless of whether any fields were filled in.
Submitting a blank form produces 100% dataset completeness and zero data.

**Data element completeness** (field-level) counts whether a specific field
was actually filled in. This is what matters for analysis — a facility that
submitted an empty form contributes nothing to the numerator or denominator of
any indicator.

DHIS2 reports dataset completeness by default, and obtaining data element
completeness requires extra steps that are not obvious in the interface.

### The WHO guidance gap

The WHO Data Quality Tool manual conflates the two levels when it states:

> *"If DHIS2 shows reliable Reporting rates for the respective data set, you can
> also use the Data Visualizer to review the trend in completeness. Such a chart
> will also tell you for which years there are data for the data element."*

This is only true if submitted forms are never blank — an assumption that fails
routinely in practice. Dataset reporting rates should not be accepted as a proxy
for data element availability.

---

## 2  Obtaining facility counts from the DHIS2 interface

The DHIS2 Data Visualizer app can display the *count* of facilities that reported
a given data element, but the path to that view is unintuitive:

1. In the Data Visualizer, build a pivot table with the data element of interest.
2. Open **Options → Data** and change *Aggregation type* from "By data element"
   to **Count**.
3. Click **Update**.

The resulting table looks identical to a value table — there is no visual
indicator that you are now seeing counts rather than sums. Furthermore, counts and
values cannot be shown side by side in the same table; each requires a separate
download. The two files must then be merged manually on org unit and period.

Because this workflow is non-obvious and inconvenient, many analysts never see
the facility counts alongside their data. The gap between expected and actual
reporting facilities goes unnoticed.

A step-by-step walkthrough of this process in the DHIS2 Sierra Leone demo
instance, with screenshots, is available in:
*Facility Counts in DHIS2*
(`../DHIS2 Best Practices/ms_body/Facility_Counts_in_DHIS2.html`).

---

## 3  What the MG2 API download returns

MG2 queries the DHIS2 analytics API at the **leaf-facility level** — the
organisational unit where data was originally entered. Because each row represents
one facility's report for one period, the COUNT returned by the API for each row
is always **1**: either the facility reported that value (COUNT = 1, SUM = the
value entered) or the row is absent.

This is a deliberate design choice. Fetching at the leaf level avoids the
ambiguity of aggregate rows, where COUNT can exceed 1 because it sums counts
across all subordinate facilities.

### The disaggregation counting problem

A data element is rarely a single field. In the Lao PDR malaria system, for
example, "MAL - RDT positive malaria cases" is collected for six age-by-sex
combinations (0–4 years, 5–14 years, 15+ years × Female, Male). Each combination
is a separate row in the API response. The demo file below shows this structure
for five Savannakhet facilities in Q1 2024.

```{r raw-structure}
library(MG2)
library(dplyr)

demo  <- mg2_demo_dir()
raw   <- readRDS(file.path(demo, "reporting_raw_demo.rds"))

# Show the first facility, one month, one element
raw |>
  filter(
    orgUnitName == "HC Asing",
    format(Month) == "2024 Jan",
    dataElement  == "MAL - RDT positive malaria cases"
  ) |>
  select(orgUnitName, dataElement, Category, Month, COUNT, SUM) |>
  knitr::kable(caption = "Six rows returned for one facility, one element, one month")
```

Every COUNT is 1. SUM holds the reported value. The facility clearly reported
this element in January 2024 — but how should we count it?

If we sum the COUNT column across all six categories we get 6, not 1. This
inflated count is exactly what the DHIS2 Data Visualizer returns when you select
"Totals" rather than individual disaggregations — it adds up the counts from all
category rows for that element.

---

## 4  Two rules for resolving the disaggregation ambiguity

MG2 offers two rules for deciding whether a facility "reported" an element in a
given month, controlled by the `reporting_rule` argument to
`mostFrequentReportingOUs()` (and the corresponding UI option in the Reporting
tab):

| Rule | Behaviour | When to use |
|---|---|---|
| `all_elements` — *at least one category* | A facility counts as reporting if **any** disaggregation (age/sex group) has a non-missing value | When the aggregate total is what you care about; partial age-group reporting is acceptable |
| `all_categories` — *every category present* | A facility counts as reporting only if **all** disaggregations have non-missing values | When you need a complete breakdown for age- or sex-specific rates |

The stricter rule (`all_categories`) always produces fewer or equal champion
facilities than `all_elements`.

### Demonstration

```{r count-by-rule}
library(tsibble)

# Facilities that reported RDT positive in Jan 2024, by rule
rdt_jan <- raw |>
  filter(
    dataElement  == "MAL - RDT positive malaria cases",
    format(Month) == "2024 Jan"
  )

# all_elements: any category present
any_cat <- rdt_jan |>
  filter(!is.na(SUM)) |>
  distinct(orgUnitName) |>
  nrow()

# all_categories: all 6 sex×age categories present
all_cat <- rdt_jan |>
  filter(!is.na(SUM)) |>
  group_by(orgUnitName) |>
  summarise(n_cats = n(), .groups = "drop") |>
  filter(n_cats == 6) |>
  nrow()

cat(sprintf(
  "Jan 2024, RDT positive (5 facilities shown):\n  Any category present : %d\n  All categories present: %d\n",
  any_cat, all_cat
))
```

```{r count-table, echo=FALSE}
# Summary table across all three months and both elements
summary_tbl <- raw |>
  filter(dataElement == "MAL - RDT positive malaria cases") |>
  filter(!is.na(SUM)) |>
  group_by(orgUnitName, Month) |>
  summarise(
    n_cats_reported = n(),
    total_cats      = 6L,   # 3 age × 2 sex
    any_rule        = n_cats_reported >= 1L,
    all_rule        = n_cats_reported == total_cats,
    .groups = "drop"
  )

result <- summary_tbl |>
  group_by(Month) |>
  summarise(
    `Any category` = sum(any_rule),
    `All categories` = sum(all_rule),
    .groups = "drop"
  ) |>
  mutate(Month = format(Month))

knitr::kable(
  result,
  caption = paste0(
    "Facilities reporting MAL - RDT positive malaria cases under each rule,",
    " 5 Savannakhet facilities (HC Asing, HC Densavan, HC Denvilai, ",
    "HC Dongnasan, 1305 DH Xepon)"
  )
)
```

In this small sample every facility reported all six categories, so both rules
produce the same count. In a full national dataset, some facilities omit one or
two age groups — the two rules then diverge, and the choice meaningfully affects
how many facilities qualify as consistently reporting.

---

## 5  Suspected ≠ Tested

One important consequence of disaggregation structure is that related elements
can look superficially similar but measure different things. In the Lao PDR
system:

- **MAL - Suspected malaria cases** counts patients presenting with fever who
  *may* have received a diagnostic test — a clinical tally.
- **MAL - Malaria suspects tested (RDT)** counts patients who *actually received*
  an RDT — a laboratory tally.

These are entered in different fields, from different datasets, and the number of
facilities that report each element differs substantially.

```{r suspected-vs-tested}
# Categories present in the raw demo
raw |>
  distinct(dataElement, Category) |>
  arrange(dataElement, Category) |>
  knitr::kable(caption = "Data elements and categories in the demo raw file")
```

The suspected cases element uses age-only categories (no sex split), while RDT
positive uses age × sex. This structural difference means that even the
`all_categories` rule is comparing apples to oranges if you mix these elements in
a champion criterion.

---

## 6  Summary

| Concept | Key point |
|---|---|
| Dataset completeness | Measures form submission, not field completion |
| Data element completeness | Measures whether the field of interest was filled in |
| DHIS2 interface | Counts are obtainable but hidden behind non-obvious options; cannot be shown alongside values |
| MG2 API | Leaf-level download; COUNT = 1 per row; SUM = reported value |
| Disaggregation ambiguity | Multiple rows per element per facility per month; "any category" vs "all categories" rule |
| Suspected vs. tested | Different elements, different datasets, different facility sets |

These mechanics underpin the champion-facility approach described in the companion
vignette, *Champion Facilities and Reporting Bias*, which uses MG2's Reporting
tab to identify consistently reporting facilities and adjust burden estimates for
reporting bias.
